---
title: "Hach and YSI Data Wrangling"
collaborators: Renee Grambihler
project: OceanWaterQuality
institution: Biosphere 2 Ocean
output: html_notebook
---
## Import Libraries
```{r Import Libraries}
library(tidyverse)
library(lubridate)
options(scipen = 999)
```

## Import Data
```{r import data}
# use 02-YSI-Data-QCCutoff.csv
dfYSI <- read_csv("Data/Clean/YSI/02_QualityControl/02-YSI-Data-QCCutoff.csv")

# use 02-Hach-Data-Clean-QC.csv
dfHach <- read_csv("Data/Clean/Hach/02_QualityControl/02-Hach-Data-Clean-QC.csv")

```

# Data Formatting
## YSI
### Daily Medians
YSI Exo3 Sonde records data every 15 minutes, which means there are multiple data points for each day. To use this data with the Hach data, the median value for each day should be used. Hach data is taken once per day (Mondays and Fridays usually)
```{r data format: ysi daily medians}
dfYSI_daily <- dfYSI %>%
  # create year, month, week columns
  mutate(year = year(date)) %>%
  mutate(month = month(date)) %>%
 # mutate(week = week(date)) %>%
  mutate(day = day(date)) %>%
  # group by year, month, day to take comparisons of each week within the time series
  group_by(year, month, day) %>%
  # take daily medians 
  mutate(temp_f = median(temp_f, na.rm = TRUE)) %>%
  mutate(temp_c = median(temp_c, na.rm = TRUE)) %>%
  mutate(pH_ysi = median(pH, na.rm = TRUE)) %>%
  mutate(pH_mV = median(pH_mV, na.rm = TRUE)) %>%
  mutate(salinity_psu = median(salinity_psu, na.rm = TRUE)) %>%
  mutate(conductivity_uScm = median(conductivity_uScm, na.rm = TRUE)) %>%
  mutate(nlf_conductivity_uScm = median(nlf_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(sp_conductivity_uScm = median(sp_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(do_mgL = median(do_mgL, na.rm = TRUE)) %>%
  mutate(do_saturated_percent = median(do_saturated_percent, na.rm = TRUE)) %>%
  mutate(do_local_percent = median(do_local_percent, na.rm = TRUE)) %>%
  mutate(orp_mV = median(orp_mV, na.rm = TRUE)) %>%
  mutate(chlorophyll_rfu = median(chlorophyll_rfu, na.rm = TRUE)) %>%
  mutate(total_algae_rfu = median(total_algae_rfu, na.rm = TRUE)) %>%
  mutate(total_dissolved_solids_mgL = median(total_dissolved_solids_mgL, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-time, -date, -chlorophyll_ugL, -total_algae_cellsL, -pH) %>%
  distinct() %>%
  mutate(Date = paste(year, month, day, sep = "-")) %>%
  ungroup() %>%
  select(-year, -month, -day) %>%
  relocate(Date, .before = temp_f)

```

### Weekly Medians
```{r data format: ysi weekly medians}
dfYSI_weekly <- dfYSI %>%
  # create year, month, week columns
  mutate(year = year(date)) %>%
  mutate(month = month(date)) %>%
  mutate(week = week(date)) %>%
  # group by year, month, week
  group_by(year, month, week) %>%
  # take weekly medians 
  mutate(temp_f = median(temp_f, na.rm = TRUE)) %>%
  mutate(temp_c = median(temp_c, na.rm = TRUE)) %>%
  mutate(pH_ysi = median(pH, na.rm = TRUE)) %>%
  mutate(pH_mV = median(pH_mV, na.rm = TRUE)) %>%
  mutate(salinity_psu = median(salinity_psu, na.rm = TRUE)) %>%
  mutate(conductivity_uScm = median(conductivity_uScm, na.rm = TRUE)) %>%
  mutate(nlf_conductivity_uScm = median(nlf_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(sp_conductivity_uScm = median(sp_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(do_mgL = median(do_mgL, na.rm = TRUE)) %>%
  mutate(do_saturated_percent = median(do_saturated_percent, na.rm = TRUE)) %>%
  mutate(do_local_percent = median(do_local_percent, na.rm = TRUE)) %>%
  mutate(orp_mV = median(orp_mV, na.rm = TRUE)) %>%
  mutate(chlorophyll_rfu = median(chlorophyll_rfu, na.rm = TRUE)) %>%
  mutate(total_algae_rfu = median(total_algae_rfu, na.rm = TRUE)) %>%
  mutate(total_dissolved_solids_mgL = median(total_dissolved_solids_mgL, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-time, -date, -chlorophyll_ugL, -total_algae_cellsL, -pH) %>%
  distinct() %>%
  mutate(Date = paste(year, month, week, sep = "-")) %>%
  ungroup() %>%
  select(-year, -month, -week) %>%
  relocate(Date, .before = temp_f)

```

### Monthly Medians
```{r data format: ysi monthly medians}
dfYSI_monthly <- dfYSI %>%
  # create year, month columns
  mutate(year = year(date)) %>%
  mutate(month = month(date)) %>%
  # group by year, month
  group_by(year, month) %>%
  # take monthly medians 
  mutate(temp_f = median(temp_f, na.rm = TRUE)) %>%
  mutate(temp_c = median(temp_c, na.rm = TRUE)) %>%
  mutate(pH_ysi = median(pH, na.rm = TRUE)) %>%
  mutate(pH_mV = median(pH_mV, na.rm = TRUE)) %>%
  mutate(salinity_psu = median(salinity_psu, na.rm = TRUE)) %>%
  mutate(conductivity_uScm = median(conductivity_uScm, na.rm = TRUE)) %>%
  mutate(nlf_conductivity_uScm = median(nlf_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(sp_conductivity_uScm = median(sp_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(do_mgL = median(do_mgL, na.rm = TRUE)) %>%
  mutate(do_saturated_percent = median(do_saturated_percent, na.rm = TRUE)) %>%
  mutate(do_local_percent = median(do_local_percent, na.rm = TRUE)) %>%
  mutate(orp_mV = median(orp_mV, na.rm = TRUE)) %>%
  mutate(chlorophyll_rfu = median(chlorophyll_rfu, na.rm = TRUE)) %>%
  mutate(total_algae_rfu = median(total_algae_rfu, na.rm = TRUE)) %>%
  mutate(total_dissolved_solids_mgL = median(total_dissolved_solids_mgL, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-time, -date, -chlorophyll_ugL, -total_algae_cellsL, -pH) %>%
  distinct() %>%
  mutate(Date = paste(year, month, sep = "-")) %>%
  ungroup() %>%
  select(-year, -month) %>%
  relocate(Date, .before = temp_f)

```

### Yearly Medians
```{r data format: ysi yearly medians}
dfYSI_yearly <- dfYSI %>%
  # create year, month columns
  mutate(year = year(date)) %>%
  # group by year
  group_by(year) %>%
  # take yearly medians 
  mutate(temp_f = median(temp_f, na.rm = TRUE)) %>%
  mutate(temp_c = median(temp_c, na.rm = TRUE)) %>%
  mutate(pH_ysi = median(pH, na.rm = TRUE)) %>%
  mutate(pH_mV = median(pH_mV, na.rm = TRUE)) %>%
  mutate(salinity_psu = median(salinity_psu, na.rm = TRUE)) %>%
  mutate(conductivity_uScm = median(conductivity_uScm, na.rm = TRUE)) %>%
  mutate(nlf_conductivity_uScm = median(nlf_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(sp_conductivity_uScm = median(sp_conductivity_uScm, na.rm = TRUE)) %>%
  mutate(do_mgL = median(do_mgL, na.rm = TRUE)) %>%
  mutate(do_saturated_percent = median(do_saturated_percent, na.rm = TRUE)) %>%
  mutate(do_local_percent = median(do_local_percent, na.rm = TRUE)) %>%
  mutate(orp_mV = median(orp_mV, na.rm = TRUE)) %>%
  mutate(chlorophyll_rfu = median(chlorophyll_rfu, na.rm = TRUE)) %>%
  mutate(total_algae_rfu = median(total_algae_rfu, na.rm = TRUE)) %>%
  mutate(total_dissolved_solids_mgL = median(total_dissolved_solids_mgL, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-time, -date, -chlorophyll_ugL, -total_algae_cellsL, -pH) %>%
  distinct() %>%
  mutate(Date = paste(year, sep = "-")) %>%
  ungroup() %>%
  select(-year) %>%
  relocate(Date, .before = temp_f)

```


## Hach
```{r data format: Hach daily values}
dfHach_daily <- dfHach %>%
  filter(Location == "Ocean") %>%
  select(-Location) %>%
  rename(pH_Hach = pH) %>%
  mutate(Date = as.character(Date))

```


### Weekly Medians
```{r data format: Hach weekly medians}
dfHach_weekly <- dfHach_daily %>%
  # create year, month, week columns
  mutate(year = year(Date)) %>%
  mutate(month = month(Date)) %>%
  mutate(week = week(Date)) %>%
  # group by year, month, week to take comparisons of each week within the time series
  group_by(year, month, week) %>%
  # take weekly medians since some data is taking 2x per week
  mutate(Turbidity = median(Turbidity, na.rm = TRUE)) %>%
  mutate(Phosphate = median(Phosphate, na.rm = TRUE)) %>%
  mutate(`Nitrate MR` = median(`Nitrate MR`, na.rm = TRUE)) %>%
  mutate(`Nitrate HR` = median(`Nitrate HR`, na.rm = TRUE)) %>%
  mutate(pH_Hach = median(pH_Hach, na.rm = TRUE)) %>%
  mutate(Alkalinity = median(Alkalinity, na.rm = TRUE)) %>%
  mutate(Silica = median(Silica, na.rm = TRUE)) %>%
  mutate(Iron = median(Iron, na.rm = TRUE)) %>%
  select(-Date) %>%
  distinct() %>%
  mutate(Date = paste(year, month, week, sep = "-")) %>%
  ungroup() %>%
  select(-year, -month, -week) %>%
  relocate(Date, .before = Turbidity)


```

### Monthly Medians
```{r data format: Hach monthly medians}
dfHach_monthly <- dfHach_daily %>%
  # create year, month columns
  mutate(year = year(Date)) %>%
  mutate(month = month(Date)) %>%
  # group by year, month to take comparisons of each month within the time series
  group_by(year, month) %>%
  # take monthly medians
  mutate(Turbidity = median(Turbidity, na.rm = TRUE)) %>%
  mutate(Phosphate = median(Phosphate, na.rm = TRUE)) %>%
  mutate(`Nitrate MR` = median(`Nitrate MR`, na.rm = TRUE)) %>%
  mutate(`Nitrate HR` = median(`Nitrate HR`, na.rm = TRUE)) %>%
  mutate(pH_Hach = median(pH_Hach, na.rm = TRUE)) %>%
  mutate(Alkalinity = median(Alkalinity, na.rm = TRUE)) %>%
  mutate(Silica = median(Silica, na.rm = TRUE)) %>%
  mutate(Iron = median(Iron, na.rm = TRUE)) %>%
  select(-Date) %>%
  distinct() %>%
  mutate(Date = paste(year, month, sep = "-")) %>%
  ungroup() %>%
  select(-year, -month) %>%
  relocate(Date, .before = Turbidity)


```

### Yearly Medians
```{r data format: Hach yearly medians}
dfHach_yearly <- dfHach_daily %>%
  # create year column
  mutate(year = year(Date)) %>%
  # group by year to take comparisons of each year within the time series
  group_by(year) %>%
  # take yearly medians
  mutate(Turbidity = median(Turbidity, na.rm = TRUE)) %>%
  mutate(Phosphate = median(Phosphate, na.rm = TRUE)) %>%
  mutate(`Nitrate MR` = median(`Nitrate MR`, na.rm = TRUE)) %>%
  mutate(`Nitrate HR` = median(`Nitrate HR`, na.rm = TRUE)) %>%
  mutate(pH_Hach = median(pH_Hach, na.rm = TRUE)) %>%
  mutate(Alkalinity = median(Alkalinity, na.rm = TRUE)) %>%
  mutate(Silica = median(Silica, na.rm = TRUE)) %>%
  mutate(Iron = median(Iron, na.rm = TRUE)) %>%
  select(-Date) %>%
  distinct() %>%
  mutate(Date = paste(year, sep = "-")) %>%
  ungroup() %>%
  select(-year) %>%
  relocate(Date, .before = Turbidity)


```


## Combine Data
### Daily
```{r data format: combine daily data}
# combine dfYSI_daily and dfHach_daily, with dfYSI_daily being first because it has more observations
dfYSIHach_daily <- left_join(dfYSI_daily, dfHach_daily) %>%
  rename(`Chlorophyll (RFU)` = chlorophyll_rfu,
         `Conductivity (µS/cm)` = conductivity_uScm,
         `Dissolved Oxygen, Local (%)` = do_local_percent,
         `Dissolved Oxygen (mg/L)` = do_mgL,
         `Dissolved Oxygen, Saturated (%)` = do_saturated_percent,
         `Conductivity, Non-Linear Function (µS/cm)` = nlf_conductivity_uScm,
         `Oxygen Reduction Potential (mV)` = orp_mV,
         `pH (mV)` = pH_mV,
         `pH (Hach)` = pH_Hach,
         `pH (YSI)` = pH_ysi,
         `Salinity (PSU)` = salinity_psu,
         `Conductivity, Specific (µS/cm)` = sp_conductivity_uScm,
         `Temperature (ºC)` = temp_c,
         `Temperature (ºF)` = temp_f,
         `Total Algae (RFU)` = total_algae_rfu,
         `Total Dissolved Solids (mg/L)` = total_dissolved_solids_mgL)

```

### Weekly
```{r data format: combine weekly data}
# combine dfYSI_weekly and dfHach_weekly, with dfYSI_weekly being first because it has more observations
dfYSIHach_weekly <- left_join(dfYSI_weekly, dfHach_weekly) %>%
  rename(`Chlorophyll (RFU)` = chlorophyll_rfu,
         `Conductivity (µS/cm)` = conductivity_uScm,
         `Dissolved Oxygen, Local (%)` = do_local_percent,
         `Dissolved Oxygen (mg/L)` = do_mgL,
         `Dissolved Oxygen, Saturated (%)` = do_saturated_percent,
         `Conductivity, Non-Linear Function (µS/cm)` = nlf_conductivity_uScm,
         `Oxygen Reduction Potential (mV)` = orp_mV,
         `pH (mV)` = pH_mV,
         `pH (Hach)` = pH_Hach,
         `pH (YSI)` = pH_ysi,
         `Salinity (PSU)` = salinity_psu,
         `Conductivity, Specific (µS/cm)` = sp_conductivity_uScm,
         `Temperature (ºC)` = temp_c,
         `Temperature (ºF)` = temp_f,
         `Total Algae (RFU)` = total_algae_rfu,
         `Total Dissolved Solids (mg/L)` = total_dissolved_solids_mgL)

```

### Monthly
```{r data format: combine monthly data}
# combine dfYSI_monthly and dfHach_monthly, with dfYSI_monthly being first because it has more observations
dfYSIHach_monthly <- left_join(dfYSI_monthly, dfHach_monthly) %>%
  rename(`Chlorophyll (RFU)` = chlorophyll_rfu,
         `Conductivity (µS/cm)` = conductivity_uScm,
         `Dissolved Oxygen, Local (%)` = do_local_percent,
         `Dissolved Oxygen (mg/L)` = do_mgL,
         `Dissolved Oxygen, Saturated (%)` = do_saturated_percent,
         `Conductivity, Non-Linear Function (µS/cm)` = nlf_conductivity_uScm,
         `Oxygen Reduction Potential (mV)` = orp_mV,
         `pH (mV)` = pH_mV,
         `pH (Hach)` = pH_Hach,
         `pH (YSI)` = pH_ysi,
         `Salinity (PSU)` = salinity_psu,
         `Conductivity, Specific (µS/cm)` = sp_conductivity_uScm,
         `Temperature (ºC)` = temp_c,
         `Temperature (ºF)` = temp_f,
         `Total Algae (RFU)` = total_algae_rfu,
         `Total Dissolved Solids (mg/L)` = total_dissolved_solids_mgL)

```

### Yearly
```{r data format: combine yearly data}
# combine dfYSI_yearly and dfHach_yearly, with dfYSI_yearly being first because it has more observations
dfYSIHach_yearly <- left_join(dfYSI_yearly, dfHach_yearly) %>%
  rename(`Chlorophyll (RFU)` = chlorophyll_rfu,
         `Conductivity (µS/cm)` = conductivity_uScm,
         `Dissolved Oxygen, Local (%)` = do_local_percent,
         `Dissolved Oxygen (mg/L)` = do_mgL,
         `Dissolved Oxygen, Saturated (%)` = do_saturated_percent,
         `Conductivity, Non-Linear Function (µS/cm)` = nlf_conductivity_uScm,
         `Oxygen Reduction Potential (mV)` = orp_mV,
         `pH (mV)` = pH_mV,
         `pH (Hach)` = pH_Hach,
         `pH (YSI)` = pH_ysi,
         `Salinity (PSU)` = salinity_psu,
         `Conductivity, Specific (µS/cm)` = sp_conductivity_uScm,
         `Temperature (ºC)` = temp_c,
         `Temperature (ºF)` = temp_f,
         `Total Algae (RFU)` = total_algae_rfu,
         `Total Dissolved Solids (mg/L)` = total_dissolved_solids_mgL)

```


# Write Data Files
```{r write csv file}
# create csv file of YSI and Hach daily values
write_csv(dfYSIHach_daily, 
          file = "Data/Clean/HachYSI/01_DataWrangling/01-HachYSI-Data-Daily.csv", 
          col_names = TRUE)

# create csv file of YSI and Hach weekly values
write_csv(dfYSIHach_weekly,
          file = "Data/Clean/HachYSI/01_DataWrangling/02-HachYSI-Data-Weekly.csv",
          col_names = TRUE)

# create csv file of YSI and Hach monthly values
write_csv(dfYSIHach_monthly,
          file = "Data/Clean/HachYSI/01_DataWrangling/03-HachYSI-Data-Monthly.csv",
          col_names = TRUE)

# create csv file of YSI and Hach yearly values
write_csv(dfYSIHach_yearly,
          file = "Data/Clean/HachYSI/01_DataWrangling/04-HachYSI-Data-Yearly.csv",
          col_names = TRUE)

```

