---
title: "HACH Data Quality Control"
Collaborators: Renee Grambihler
Project: "The Biosphere 2 Ocean mesocosm: (re)engineered for state-of-the-art coral reef resilience experiments"
Institution: Biosphere 2 Reef Lab
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
# Hach Data Quality Control  
## Input and Output Files  
### Input Files  
InputData/Clean/01-DataWrangling: 
  `Hach-Data-Clean.csv`: cleaned Hach nutrient water quality data for all locations from 2017-2024  
    - variables: 
      `Date`: date in mm/dd/yy  
      `Location`: character; location of measurements; Ocean, Mangroves, Raceway 1, or Raceway 2  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
      `pH`: pH in pH units  
      `Alkalinity (mg/L CaCO3)`: numeric; calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; iron concentration in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  
      `ORP`:  numeric;oxidation-reduction potential (ORP) measured in miliVolts  
      `Initials`: character; initials of person or people taking measurements  
      `Notes`: character; any notes from that measurement  

### Output Files
InputData/Clean/02-QualityControl:
  `Hach-Data-QC.csv`: Hach nutrient water quality data after being run through quality control parameters
    - variables:  
      `Date`: date in mm/dd/yy  
      `Location`: character; location of measurements; Ocean, Mangroves, Raceway 1, or Raceway 2  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
      `pH`: pH in pH units  
      `Alkalinity (mg/L CaCO3)`: numeric; calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; iron concentration in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  


```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
# remove scientific notation
options(scipen = 999)
```

# Import Data
```{r import data}
# use B2ONutrientData-Clean.csv
# drop ORP, Initials, Notes columns
df <- read_csv("InputData/Clean/01-DataWrangling/Hach-Data-Clean.csv") %>%
  select(-ORP, -Initials, -Notes)
```

# Quality Control
For each nutrient in each location, turn values outside 2-sigma from mean into NA
```{r QC}
# make empty dataframe to store output from loop
# creating with just 2 columns for now since "date" and "location" columns will not change
dfQC <- data.frame(matrix(ncol = ncol(df)))
colnames(dfQC) <- colnames(df)

# getting list of column names from df, excluding "date" and "location" so we can use it to loop through
nutrient_list <- colnames(df)
nutrient_list <- nutrient_list[! nutrient_list %in% c("Date", "Location")]
# list of locations
location_list <- c(unique(df$Location))

for (l in location_list) {
  # filter df by location, l
  dfLocation <- df %>%
    filter(Location == l)
  
  dfLocationQC <- dfLocation %>%
    select(Date, Location)
  
  for (n in nutrient_list) {
    # select column from df to use in calculations (represented by n in the loop) and store as nutrient
    nutrient_col <- dfLocation[[n]]
    # frame with just date, location, and nutrient columns - all from df
    dfLocationNutrient <- dfLocation[, c("Date", "Location")]
    dfLocationNutrient <- cbind(dfLocationNutrient, nutrient_col)
    
    # find mean
    mean_nutrient_value <- mean(nutrient_col, na.rm = TRUE)
    # find standard deviation
    sd_nutrient_value <- sd(nutrient_col, na.rm = TRUE)
    
    # if standard deviation is < 0 or NA, make sd_nutrient_value 1 so that we don't divide by 0 in next step
    if (sd_nutrient_value < 0 | is.na(sd_nutrient_value) == TRUE){
      sd_nutrient_value <- 1
    }
    
    # go through nutrient_col; if value is NA or absolute value of z-score is > 2, make value NA, otherwise, leave alone
    dfQCLocationNutrient <- dfLocationNutrient %>%
      mutate(nutrient_col = ifelse(is.na(mean_nutrient_value) == TRUE |
                                     abs((nutrient_col-mean_nutrient_value)/sd_nutrient_value) > 2,
                                   NA,
                                   nutrient_col))
    
    # select nutrient_col as nutrient_col_qc (for redundancy)
    nutrient_col_qc <- dfQCLocationNutrient$nutrient_col
    
    # stick new QC'd nutrient column into dfLocationQC
    dfLocationQC <- dfLocationQC %>%
      cbind(nutrient_col_qc)

    # set the column name of dfLocationQC that is currently called "nutrient" to be what n currently represents in the loop. this is a really important step!!!!!
    colnames(dfLocationQC)[colnames(dfLocationQC) == "nutrient_col_qc"] = n
  }
  
  # bind all rows and columns of dfLocationQC to dfQC
  dfQC <- dfQC %>%
    rbind(dfLocationQC)
  
}

# remove rows where Date is NA; make sure Date values are in yyyy-mm-dd format and not numeric
dfQC <- dfQC %>%
  filter(is.na(Date) == FALSE) %>%
  mutate(Date = as.Date(Date))

```

### Save QC'd Data
```{r save data}
write_csv(dfQC,
          file = "InputData/Clean/02-QualityControl/Hach-Data-QC.csv")
```

