---
title: "Ocean Data Wrangling: 2011-2022"
Collaborators: Renee Grambihler
Project: OceanWaterQuality
Institution: Biosphere 2 Ocean
output: html_notebook
---
## Import Libraries
```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
library(here)
# turns off scientific notation
options(scipen = 999)
```

# Import All Data Files
## SCADA Files from Wei-Ren: 2011 - 2021
This combines some but not all files from SCADA_2011_2021 folder which originally came from Wei-Ren Ng, Dr. Matej Durcik, and previously used SCADA databases. See 'YSI-metadata.xlsx' file for more details on origin. The files used below have the most complete data for the time period of 2011-2021.

Code here adds columns with NAs to some files that are present with actual data in other files, and it does some basic filtering of obvious data-entry or instrument errors.

```{r import-SCADA-files-from-2011-2021}
# file path to folder containing old scada database 2011-2021 ocean water quality files
path <- "Data/Raw/YSI/SCADA/SCADA_2011_2021"

# data from 7/22/11 - 8/4/14
ocean_data_2011_2014 <- read_csv(here(path, "Ocean_FTP_Data - pre 2014-8-11.csv"),
                                 skip = 1,
                                 col_names = c("datetime",
                                               "Temperature (ºF)",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (µS/cm)",
                                               "Oxygen Reduction Potential (mV)")) %>%
  # add a column called chlorophyll_ugl after `Oxygen Reduction Potential (mV)` and fill it with NA values
  add_column(`Chlorophyll (µg/L)` = NA,
             .after = 7)%>%
  # get rid of probable data entry errors: salinity < 1, temp < 10, pH < 0, conductivity < 2, orp < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 10, NA, `Temperature (ºC)`)) %>%
  mutate(pH = ifelse(pH < 0, NA, pH)) %>%
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` < 2, NA, `Conductivity (µS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`))

# data from 10/3/14 - 5/8/18
ocean_data_2014_2018 <- read_csv(here(path, "Ocean_FTP_Data.csv"), 
                                 skip = 1,
                                 col_names = c("datetime",
                                               "Temperature (ºF)",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (µS/cm)",
                                               "Oxygen Reduction Potential (mV)")) %>%
  # add a column called chlorophyll_ugl after `Oxygen Reduction Potential (mV)` and fill it with NA values
  add_column(`Chlorophyll (µg/L)` = NA,
             .after = 7)%>%
  # change datetime format to match previous files
  mutate(datetime = format(as.POSIXct(datetime, format = "%m/%d/%y %H:%M"), "%Y/%m/%d %H:%M:%S"))%>%
  # data in this file is from YSI not in use after 5/8/18 at 8:00am. just use data before that date
  filter(datetime <= "2018/05/08 08:00:00")%>%
  # get rid of probable data entry errors: salinity < 1, temp < 10F and 1C, pH < 3, conductivity < 2, orp < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºF)` = ifelse(`Temperature (ºF)` < 10, NA, `Temperature (ºF)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 1, NA, `Temperature (ºC)`))%>%
  mutate(pH = ifelse(pH < 3, NA, pH))%>%
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` < 2, NA, `Conductivity (µS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`))

# data from 5/8/18 - 6/12/18
ocean_data_2016_2018 <- read_csv(here(path, "Ocean_FTP_Data_Extended.csv"), 
                                 skip = 1,
                                 col_names = c("datetime",
                                               "Temperature (ºF)",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (µS/cm)",
                                               "Oxygen Reduction Potential (mV)",
                                               "Total Algae (RFU)",
                                               "Chlorophyll (µg/L)",
                                               "ro_flow_gpm")) %>%
  # don't need RO Flow Rate into Ocean. This is also only file with this data
  select(-ro_flow_gpm) %>%
  # use data after 5/8/18 at 8:00am to keep from overlapping with previous file
  filter(datetime >= "2018/05/08 08:00:00") %>%
  # get rid of probable data entry errors: salinity < 1, temp < 10, pH < 0, conductivity < 2, orp < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºF)` = ifelse(`Temperature (ºF)` < 10, NA, `Temperature (ºF)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 1, NA, `Temperature (ºC)`))%>%
  mutate(pH = ifelse(pH < 3, NA, pH))%>%
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` < 2, NA, `Conductivity (µS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`))


# data from 10/12/18 - 5/25/21
ocean_data_2018_2021 <- read_csv(here(path, "Ocean_Data_new.csv"), 
                                 skip = 1, 
                                 col_names = c("datetime",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (µS/cm)",
                                               "Oxygen Reduction Potential (mV)",
                                               "Total Algae (cells/L)",
                                               "Chlorophyll (µg/L)",
                                               "Dissolved Oxygen, Saturated (%)")) %>%
  # add temp_f column with NA values
  add_column(`Temperature (ºF)` = NA, .after = 1)%>%
  # get rid of probable data entry errors: salinity < 1, temp < 10, temp_c < 1, pH < 0, conductivity < 2, orp < 0, total algae cells < 0, chlorophyll < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºF)` = ifelse(`Temperature (ºF)` < 10, NA, `Temperature (ºF)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 1, NA, `Temperature (ºC)`))%>%
  mutate(pH = ifelse(pH < 3, NA, pH))%>%
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` < 2, NA, `Conductivity (µS/cm)`))%>%
  # ~3 days in January 2019 were reading `Conductivity (µS/cm)` in the 10,000s instead of 10s, so there must have been a conversion issue for those days
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` > 1000, `Conductivity (µS/cm)`/1000, `Conductivity (µS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`))%>%
  mutate(`Total Algae (cells/L)` = ifelse(`Total Algae (cells/L)` < 0, NA, `Total Algae (cells/L)`))%>%
  mutate(`Chlorophyll (µg/L)` = ifelse(`Chlorophyll (µg/L)` < 0, NA, `Chlorophyll (µg/L)`))

```


## Old B2 SCADA database files (10/21/19 - 2/23/22): old YSI sonde and new YSI EXO3 sonde
These files are found in the SCADA_2019_2022 folder. They originate from the old B2 SCADA database which is no longer available on biosphere2.org. Some of the data will overlap with the SCADA data from Wei-Ren, but I want to make sure we have absolutely everything possible; will take out repeated data later
### Old YSI Sonde (10/21/19-5/26/21)
```{r import-files-old-db-ysi}
# file path to folder containing old scada database 10/21/19-2/23/22 ocean water quality files
path <- "Data/Raw/YSI/SCADA/SCADA_2019_2022"

# total algae phycoerythrin in cells/L 
ysi_total_algae <- read_csv(here(path,
                                       "UAB2_B240001_YSI_OCEANBGA_PE_593908.csv"),
                                  # select columns with date, time, and the data values
                                    col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Total Algae (cells/L)` = VALUE)

# chlorophyll in ug/L 
ysi_chlorophyll <- read_csv(here(path,
                                       "UAB2_B240001_YSI_OCEANCHLORO_595563.csv"),
                                  # select columns with date, time, and the data values
                                  col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Chlorophyll (µg/L)` = VALUE)

# electrical conductivity in mS/cm 
ysi_conductivity <- read_csv(here(path, 
                                        "UAB2_B240001_YSI_OCEANCONDUC_597443.csv"),
                                   # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE"))%>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Conductivity (µS/cm)` = VALUE)

# reduction potential (ORP) in mV 
ysi_orp <- read_csv(here(path, "UAB2_B240001_YSI_OCEANORP_598894.csv"),
                          # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Oxygen Reduction Potential (mV)` = VALUE)

# pH in pH units 
ysi_pH <- read_csv(here(path, "UAB2_B240001_YSI_OCEANPH_600334.csv"),
                         # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         pH = VALUE)

# salinity in ppt 
ysi_salinity <- read_csv(here(path, "UAB2_B240001_YSI_OCEANSALANI_602884.csv"),
                               # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Salinity (PSU)` = VALUE)

# temperature in degrees Celcius 
ysi_temp_c <- read_csv(here(path, "UAB2_B240001_YSI_OCEANTMP_604270.csv"),
                             # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Temperature (ºC)` = VALUE)

```

### YSI EXO 3 Sonde (5/26/21-2/23/22)
```{r import-files-old-db-ysi-exo-3}
# chlorophyll in RFU 
exo_chlorophyll <- read_csv(here(path, 
                                       "UAB2_B240001_YSI_EXO3_CHLORO_617107.csv"),
                                  # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Chlorophyll (RFU)` = VALUE)

# electrical conductivity in uS/cm 
exo_conductivity <- read_csv(here(path, 
                                        "UAB2_B240001_YSI_EXO3_COND_U_617936.csv"),
                                   # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  
  rename(datetime = TIMESTAMP,
         `Conductivity (µS/cm)` = VALUE)

# nLF conductivity in uS/cm 
exo_nlf_conductivity <- read_csv(here(path, 
                                      "UAB2_B240001_YSI_EXO3_NLF_CO_661668.csv"),
                                      # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Conductivity, Non-Linear Function (µS/cm)` = VALUE)

# dissolved oxygen in % local 
exo_do_local <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_ODO__L_632594.csv"),
                               # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen, Local (%)` = VALUE)

# dissolved oxygen in % saturated 
exo_do_saturated <- read_csv(here(path, 
                                        "UAB2_B240001_YSI_EXO3_ODO__S_635338.csv"),
                                   # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen, Saturated (%)` = VALUE)

# dissolved oxygen in mg/L 
exo_do <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_ODO_MG_645798.csv"),
                         # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen (mg/L)` = VALUE)

# reduction potential (ORP) in mV 
exo_orp <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_ORP_MV_647084.csv"),
                          # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Oxygen Reduction Potential (mV)` = VALUE)

# pH in pH units 
exo_ph <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_PH_662411.csv"),
                         # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         pH = VALUE)

# pH in mV 
exo_ph_mv <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_PH_MV_663459.csv"),
                            # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `pH (mV)` = VALUE)

# salinity in psu 
exo_salinity <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_SAL_PS_647937.csv"),
                               # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Salinity (PSU)` = VALUE)

# specific conductivity in uS/cm 
exo_sp_conductivity <- read_csv(here(path, 
                                           "UAB2_B240001_YSI_EXO3_SPCOND_648943.csv"),
                                      # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Conductivity, Specific (µS/cm)` = VALUE)

# total algae phycoerythrin in RFU 
exo_total_algae <- read_csv(here(path, 
                                       "UAB2_B240001_YSI_EXO3_TAL_PE_650659.csv"),
                                  # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Total Algae (RFU)` = VALUE)

# total dissolved solids in mg/L 
exo_total_dissolved_solids <- read_csv(here(path,
                                          "UAB2_B240001_YSI_EXO3_TDS_MG_653051.csv"),
                                          # select columns with date, time, and the data values
                                          col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Total Dissolved Solids (mg/L)` = VALUE)

# temperature in degrees Celcius 
exo_temp_c <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_TEMP_C_655738.csv"),
                             # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Temperature (ºC)` = VALUE) %>%
  # keep data taken at the quarter hours and remove other data; Sonde or database must have gotten switched to recording in 1-minute intervals for a period of time here
  filter(str_detect(datetime, pattern = ":00:") | str_detect(datetime, pattern = ":15:") | str_detect(datetime, pattern = ":30:") | str_detect(datetime, pattern = ":45:"))

# temperature in degrees Fahrenheit 
exo_temp_f <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_TEMP_F_659351.csv"),
                             # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Temperature (ºF)` = VALUE)

```


# Data Wrangling
Join all dataframes together so there is one 2011-2022 dataframe. 
## 2011-2021 SCADA Data
```{r data-wrangling-2011-2021}
# create 2011-2021 ocean data dataframe
OceanDataFrame_2011_2021 <- rbind(ocean_data_2011_2014, 
                              ocean_data_2014_2018,
                              #all columns except total_algae_rfu (column 8)
                              ocean_data_2016_2018[,c(1:7,9)],
                              ocean_data_2018_2021[,c(1:7,9)])%>%
  # join total_algae_rfu column on end
  left_join(ocean_data_2016_2018[,c(1,8)], by = join_by(datetime))%>%
  left_join(ocean_data_2018_2021[,c(1,8)], by = join_by(datetime))%>%
  left_join(ocean_data_2018_2021[,c(1,10)], by = join_by(datetime)) %>%
  # turn units into uS/cm instead of mS/cm
  mutate(`Conductivity (µS/cm)` = `Conductivity (µS/cm)`*1000)

```

## Old Database Data: 2019-2022
```{r data-wrangling-old-database}
# create old database YSI dataframe by joining the files for each variable from this instrument based on the datetime column
YsiFrame <- ysi_temp_c %>%
  left_join(ysi_pH, by = join_by(datetime)) %>%
  left_join(ysi_salinity, by = join_by(datetime)) %>%
  left_join(ysi_conductivity, by = join_by(datetime)) %>%
  left_join(ysi_orp, by = join_by(datetime)) %>%
  left_join(ysi_chlorophyll, by = join_by(datetime))%>%
  left_join(ysi_total_algae, by = join_by(datetime)) %>%
  # turn units into uS/cm instead of mS/cm
  mutate(`Conductivity (µS/cm)` = `Conductivity (µS/cm)`*1000) 


# create old database YSI EXO3 dataframe by joining the files for each variable from this instrument based on the datetime column
YsiExoFrame <- exo_temp_c %>%
  left_join(exo_ph, by = join_by(datetime)) %>%
  left_join(exo_salinity, by = join_by(datetime)) %>%
  left_join(exo_conductivity, by = join_by(datetime)) %>%
  left_join(exo_orp, by = join_by(datetime)) %>%
  left_join(exo_total_algae, by = join_by(datetime)) %>%
  left_join(exo_chlorophyll, by = join_by(datetime)) %>%
  left_join(exo_do, by = join_by(datetime)) %>%
  left_join(exo_do_local, by = join_by(datetime)) %>%
  left_join(exo_do_saturated, by = join_by(datetime)) %>%
  left_join(exo_nlf_conductivity, by = join_by(datetime)) %>%
  left_join(exo_ph_mv, by = join_by(datetime)) %>%
  left_join(exo_sp_conductivity, by = join_by(datetime)) %>%
  left_join(exo_temp_f, by = join_by(datetime)) %>%
  left_join(exo_total_dissolved_solids, by = join_by(datetime))

# create dataframe of all old database YSI Sonde data
YsiAllFrame <- rbind(YsiFrame[,1:6], YsiExoFrame[,1:6]) %>%
  left_join(YsiFrame[,c(1,7:8)], by = join_by(datetime)) %>%
  left_join(YsiExoFrame[,c(1,7:16)], by = join_by(datetime))

# edit column placement of YsiAllFrame to match up common columns with OceanDataFrame_2011_2021
YsiAllFrameEdit <- YsiAllFrame %>%
  relocate(`Temperature (ºF)`, .before = `Temperature (ºC)`) %>%
  relocate(`Total Algae (RFU)`, .after = `Chlorophyll (µg/L)`) %>%
  relocate(`Dissolved Oxygen, Saturated (%)`, .after = `Total Algae (cells/L)`)


```

## Combine Dataframes
Combine dataframes, keep unique rows, and move columns around so that similar variables are grouped together
```{r data-wrangling-combine-data}

OceanWaterQualityFrame <- rbind(OceanDataFrame_2011_2021, YsiAllFrameEdit[, 1:11]) %>%
  left_join(YsiAllFrameEdit[,c(1,12:18)], by = join_by(datetime)) %>%
  unique() %>%
  relocate(`pH (mV)`, .after = pH) %>%
  relocate(`Conductivity, Non-Linear Function (µS/cm)`, 
           `Conductivity, Specific (µS/cm)`, .after = `Conductivity (µS/cm)`) %>%
  relocate(`Dissolved Oxygen (mg/L)`, 
           `Dissolved Oxygen, Saturated (%)`, 
           `Dissolved Oxygen, Local (%)`, .after = `Conductivity, Specific (µS/cm)`) %>%
  relocate(`Chlorophyll (RFU)`, .after = `Oxygen Reduction Potential (mV)`)

```

# Write Data to File
```{r write-data-to-file}
write_csv(OceanWaterQualityFrame,
          file = "Data/Clean/YSI/01_DataWrangling/01-YSI-Data-Clean-2011-2022.csv",
          col_names = TRUE)
```



```{r eval=FALSE, echo=FALSE, import-and-format-UTF-16-files}
# The following chunks are original code from Dan (and Renee) for EXO3 files -- can be ignored
#file path for your data directory. might work with the "Box Sync" client as well, haven't tried
path16 <- "/Users/grambihler/Documents/Data/GitHub/OceanWaterQuality/YSIExoSonde/YSI-sonde-raw-data/UTF-16-versions" 

# create ExoFrame16: use path to create list of UTF-16 files to read in, read in files in path, specify column data types, skip the first 9 rows, and specify UTF-16 encoding for files from YSI exo sonde
ExoFrame16 <- list.files(path = path16,
                       pattern = ".csv") %>%
            map_df(~read_csv(.,
                             col_types = cols(.default = col_double(),
                                              `Date (MM/DD/YYYY)` = col_date(format = '%m/%d/%Y'),
                                              `Site Name` = col_character(),
                                              `Time (HH:mm:ss)` = col_time(format = "%H:%M:%S")),
                             skip = 9,
                             locale=locale(encoding="UTF-16")))
```

```{r eval=FALSE, echo=FALSE, Set-Working-Directory}
setwd("C:/Users/fried/OneDrive/Desktop/YSI Data") #your data directory. might work with the "Box Sync" client as well, haven't tried
```

```{r eval=FALSE, echo=FALSE, File-Formatting}
files <- list.files(pattern = ".csv") #create list of files in directory

#bind files into a list. skips the first 8 lines (EXO summary data). creates a column noting the source file of each entry
exolist<-Map(cbind, lapply(files, fread,sep=",",skip=8), filename = files) 

#fread function can't work with the files as they come out of YSI (in UTF-16 format). 
#I had to re-save them in Excel to change them to UTF-8. The YSI outputs should either be converted in bulk in the future, or maybe there's a setting in the KorEXO template.

```

```{r eval=FALSE, echo=FALSE, Data-Wrangling}
#bind list entries into single merged data frame
exodata <- rbindlist(exolist,use.names=TRUE,fill=TRUE) 

#rename columns to easier names
colnames(exodata)<-c("Date","Time","Time.fracsec","Sitename","Chlor.RFU","Cond.uScm",
                     "nLF.Cond.uScm","DO.percsat","DO.perclocal","DO.mgL","ORP.mV",
                     "S.PSU","SpCond.uScm","TAL.PE.RFU","TDS.mgL","Wiperpos.V",
                     "pH","pH.mV","TempC","BattV","CablepwrV","filename") 

#I find that R does best when you break up the date column into its parts and put them back together into a date
#break date into its parts
exodata <- exodata %>%
  separate(Date,c("Month","Day","Year"),sep="/") 

#break time into its parts
exodata <- exodata %>%
  separate(Time,c("Hour","Minute","Second"))

#paste them together again and make a date
exodata$Datetime <- ymd_hms(paste(exodata$Year,exodata$Month,exodata$Day,
                                  " ",exodata$Hour,":",exodata$Minute,":",exodata$Second))

#create a date column again
exodata$Date <- as_date(date(exodata$Datetime))

#remove any duplicate rows
exodata <- distinct(exodata)
#remove any NAs
exodata <- subset(exodata,is.na(exodata$Datetime)==FALSE) 

#make any columns that might be stuck as factors numeric
exodata <- exodata %>% 
  mutate_at(c(9:25), as.numeric) 

#create a time tibble
exotbl <- as_tbl_time(exodata,index=Datetime)

#order by day
exotblm<- exotbl %>% 
  arrange(ymd_hms(exotbl$Datetime)) 
```

```{r eval=FALSE, echo=FALSE, Data-Analysis-Syntax-Example}
#this syntax can be used to aggregate the data by all sorts of periods including week, hour, etc
exoday<-exotbl %>%
  collapse_by("day") %>%
  group_by(Datetime,Date)%>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

```

```{r eval=FALSE, echo=FALSE, Plot-Example}
#create plots for data points

ggplot(exoday,aes(Date,DO.mgL))+
ggplot(exoday,aes(Date,pH))+
  geom_point(cex=1,alpha=0.5)+
  ylab(expression(paste("pH")))+
  xlab(element_blank())+
  theme_classic(base_size=20)+
  theme(legend.position = "bottom")


```