---
title: "Ocean Data Wrangling: 2011-2022"
Collaborators: Renee Grambihler
Project: OceanWaterQuality
Institution: Biosphere 2 Ocean
output: html_notebook
---

# Ocean Data Wrangling: 2011-2022

## Import Libraries

```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
library(here)
library(readxl)
# turns off scientific notation
options(scipen = 999)
```

## Import All Data Files

Code for importing files will include some basic wrangling such as renaming columns, adding columns (populated with NAs) to some files that are present (with actual data) in other files, filtering out obvious data entry errors, checking datatypes, splitting any datetime columns into separate Date and Time columns, and keeping only data points recorded every 15 minutes. Much of this code is redundant, but I've found that the redundancy keeps everything as similar as possible and keeps errors made in the beginning steps from popping up in the later steps.

### SCADA Files from Wei-Ren: 2011 - 2021

This combines some but not all files from SCADA_2011_2021 folder which originally came from Wei-Ren Ng, Dr. Matej Durcik, and previously used SCADA databases. See 'YSI-metadata.xlsx' file for more details on origin. The files used below have the most complete data for the time period of 2011-2021.

```{r SCADA-files-from-2011-2021-path}
# file path to folder containing old scada database 2011-2021 ocean water quality files
path <- "Data/Raw/YSI/SCADA/SCADA_2011_2021"
```

### 7/22/11 - 8/4/14

```{r import-SCADA-files-from-2011-2014}
# data from 7/22/11 - 8/4/14
ocean_data_2011_2014 <- read_csv(here(path, "Ocean_FTP_Data - pre 2014-8-11.csv"),
                                 skip = 1,
                                 col_names = c("datetime",
                                               "Temperature (ºF)",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (mS/cm)",
                                               "Oxygen Reduction Potential (mV)")) %>%
  # add a columns with NA values to match names and places of columns we will have eventually with addition of other datafiles
  add_column(`Total Algae (RFU)` = NA, .after = 7) %>%
  add_column(`Chlorophyll (µg/L)` = NA, .after = 8) %>%
  add_column(`Total Algae (cells/L)` = NA, .after = 9) %>%
  add_column(`Dissolved Oxygen, Saturated (%)` = NA, .after = 10) %>%
  # change datetime format and data type
  mutate(datetime = gsub("/", "-", datetime))%>%
  mutate(datetime = as_datetime(datetime)) %>%
  # get rid of probable data entry errors: salinity < 1, temp < 10, pH < 0, conductivity < 2, orp < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 10, NA, `Temperature (ºC)`)) %>%
  mutate(pH = ifelse(pH < 0, NA, pH)) %>%
  mutate(`Conductivity (mS/cm)` = ifelse(`Conductivity (mS/cm)` < 2, NA, `Conductivity (mS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`)) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time)) 

```

### 8/4/2014 - 10/3/2014

```{r import-SCADA-files-from-end-2014}
# data from 8/4/2014 - 10/3/2014
ocean_data_2014 <- read_csv(here(path, "Ocean_FTP_Data_Acq.csv"), 
                                 skip = 1,
                                 col_names = c("datetime",
                                               "Temperature (ºF)",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (mS/cm)",
                                               "Oxygen Reduction Potential (mV)")) %>%
  # add a columns with NA values to match names and places of columns we will have eventually with addition of other datafiles
  add_column(`Total Algae (RFU)` = NA, .after = 7) %>%
  add_column(`Chlorophyll (µg/L)` = NA, .after = 8) %>%
  add_column(`Total Algae (cells/L)` = NA, .after = 9) %>%
  add_column(`Dissolved Oxygen, Saturated (%)` = NA, .after = 10) %>%
  # change datetime format and data type
  mutate(datetime = format(as.POSIXct(datetime, format = "%m/%d/%y %H:%M"), "%Y/%m/%d %H:%M:%S"))%>%
  mutate(datetime = gsub("/", "-", datetime))%>%
  # want data to fill in from end of last file to beginning of next file since the next file has more variables even though the dates overlap
  filter(datetime > "2014-08-04 09:30:00" & datetime < "2014-10-03 09:00:00")%>%
  # get rid of probable data entry errors: salinity < 1, temp < 10F and 1C, pH < 3, conductivity < 2, orp < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºF)` = ifelse(`Temperature (ºF)` < 10, NA, `Temperature (ºF)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 1, NA, `Temperature (ºC)`))%>%
  mutate(pH = ifelse(pH < 3, NA, pH))%>%
  mutate(`Conductivity (mS/cm)` = ifelse(`Conductivity (mS/cm)` < 2, NA, `Conductivity (mS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`)) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time)) 


```

### 10/3/14 - 12/6/2016

```{r import-SCADA-files-from-2014-2016}
# data from 10/3/14 - 12/6/2016
ocean_data_2014_2016 <- read_csv(here(path, "Ocean_FTP_Data.csv"), 
                                 skip = 1,
                                 col_names = c("datetime",
                                               "Temperature (ºF)",
                                               "Temperature (ºC)",
                                               "pH",
                                               "Salinity (PSU)",
                                               "Conductivity (mS/cm)",
                                               "Oxygen Reduction Potential (mV)")) %>%
  # add a columns with NA values to match names and places of columns we will have eventually with addition of other datafiles
  add_column(`Total Algae (RFU)` = NA, .after = 7) %>%
  add_column(`Chlorophyll (µg/L)` = NA, .after = 8) %>%
  add_column(`Total Algae (cells/L)` = NA, .after = 9) %>%
  add_column(`Dissolved Oxygen, Saturated (%)` = NA, .after = 10) %>%
  # change datetime format and data type
  mutate(datetime = format(as.POSIXct(datetime, format = "%m/%d/%y %H:%M"), "%Y/%m/%d %H:%M:%S"))%>%
  mutate(datetime = gsub("/", "-", datetime))%>%
  mutate(datetime = as_datetime(datetime)) %>%
  # want data to fill in from end of last file to beginning of next file since the next file has more variables even though the dates overlap
  filter(datetime < "2016/12/06 10:00:00")%>%
  # get rid of probable data entry errors: salinity < 1, temp < 10F and 1C, pH < 3, conductivity < 2, orp < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 1, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºF)` = ifelse(`Temperature (ºF)` < 10, NA, `Temperature (ºF)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 1, NA, `Temperature (ºC)`))%>%
  mutate(pH = ifelse(pH < 3, NA, pH))%>%
  mutate(`Conductivity (mS/cm)` = ifelse(`Conductivity (mS/cm)` < 2, NA, `Conductivity (mS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`)) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time)) 

```

### 12/6/2016 - 5/25/2021

```{r import-SCADA-files-from-2016-2021}
# data from 12/6/2016 - 5/25/2021
ocean_data_2016_2021 <- read_excel(here(path, "ocean_scada_data_2016_2021.xlsx"), 
                                   skip = 1, 
                                   col_names = c("datetime",
                                                 "Temperature (ºF)",
                                                 "Temperature (ºC)",
                                                 "pH",
                                                 "Salinity (PSU)",
                                                 "Conductivity (mS/cm)",
                                                 "Oxygen Reduction Potential (mV)",
                                                 "Total Algae (RFU)",
                                                 "Chlorophyll (µg/L)",
                                                 "ro_flow_gpm",
                                                 "Total Algae (cells/L)",
                                                 "Dissolved Oxygen, Saturated (%)"), 
                                   col_types = c("date",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric",
                                                 "numeric")) %>%
  select(-ro_flow_gpm) %>%
  # change datetime format to match previous files
  mutate(datetime = gsub("/", "-", datetime))%>%
  mutate(datetime = as_datetime(datetime)) %>%
  # get rid of probable data entry errors: salinity < 1, temp < 10, temp_c < 1, pH < 0, conductivity < 2, orp < 0, total algae cells < 0, chlorophyll < 0
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 10, NA, `Salinity (PSU)`))%>%
  mutate(`Temperature (ºF)` = ifelse(`Temperature (ºF)` < 20, NA, `Temperature (ºF)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 10, NA, `Temperature (ºC)`))%>%
  mutate(pH = ifelse(pH < 3, NA, pH))%>%
  mutate(`Conductivity (mS/cm)` = ifelse(`Conductivity (mS/cm)` < 0, NA, `Conductivity (mS/cm)`))%>%
  # ~3 days in January 2019 were reading `Conductivity (mS/cm)` in the 10,000s instead of 10s, so there must have been a conversion issue for those days
  mutate(`Conductivity (mS/cm)` = ifelse(`Conductivity (mS/cm)` > 1000, NA, `Conductivity (mS/cm)`))%>%
  mutate(`Oxygen Reduction Potential (mV)` = ifelse(`Oxygen Reduction Potential (mV)` < 0, NA, `Oxygen Reduction Potential (mV)`))%>%
  mutate(`Total Algae (cells/L)` = ifelse(`Total Algae (cells/L)` < 0, NA, `Total Algae (cells/L)`))%>%
  mutate(`Chlorophyll (µg/L)` = ifelse(`Chlorophyll (µg/L)` < 0, NA, `Chlorophyll (µg/L)`)) %>%
  mutate(`Dissolved Oxygen, Saturated (%)` = ifelse(`Dissolved Oxygen, Saturated (%)` < 10, NA, `Dissolved Oxygen, Saturated (%)`)) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))  

```

### Check Check: 2011-2021 Raw Files

Check that each of the dataframes created from the imported raw data has the expected number of data entries per day. This should be no more than 96: 4 data entries per hour x 24 hours per day = 96. This will also be important later for quality control.

The expected output will be 4 empty dataframes (called tibbles here).

```{r check-imported-data-2011-2021}
ocean_data_2011_2014 %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

ocean_data_2014 %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

ocean_data_2014_2016 %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

ocean_data_2016_2021 %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows
  
```

## Old B2 SCADA database files (5/26/21 - 2/23/22): YSI EXO3 sonde

These files are found in the SCADA_2019_2022 folder. They originate from the **old** B2 SCADA database which is no longer available on biosphere2.org.

### YSI EXO 3 Sonde (5/26/21-2/23/22)

```{r import-files-old-db-ysi-exo-3}
# file path to folder containing old scada database 10/21/19-2/23/22 ocean water quality files
path <- "Data/Raw/YSI/SCADA/SCADA_2019_2022"

# chlorophyll in RFU 
exo_chlorophyll <- read_csv(here(path, 
                                       "UAB2_B240001_YSI_EXO3_CHLORO_617107.csv"),
                                  # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Chlorophyll (RFU)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# electrical conductivity in uS/cm 
exo_conductivity <- read_csv(here(path, 
                                        "UAB2_B240001_YSI_EXO3_COND_U_617936.csv"),
                                   # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  
  rename(datetime = TIMESTAMP,
         `Conductivity (µS/cm)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# nLF conductivity in uS/cm 
exo_nlf_conductivity <- read_csv(here(path, 
                                      "UAB2_B240001_YSI_EXO3_NLF_CO_661668.csv"),
                                      # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Conductivity, Non-Linear Function (µS/cm)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# dissolved oxygen in % local 
exo_do_local <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_ODO__L_632594.csv"),
                               # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen, Local (%)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# dissolved oxygen in % saturated 
exo_do_saturated <- read_csv(here(path, 
                                        "UAB2_B240001_YSI_EXO3_ODO__S_635338.csv"),
                                   # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen, Saturated (%)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# dissolved oxygen in mg/L 
exo_do <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_ODO_MG_645798.csv"),
                         # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen (mg/L)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# reduction potential (ORP) in mV 
exo_orp <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_ORP_MV_647084.csv"),
                          # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Oxygen Reduction Potential (mV)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# pH in pH units 
exo_ph <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_PH_662411.csv"),
                         # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         pH = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# pH in mV 
exo_ph_mv <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_PH_MV_663459.csv"),
                            # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `pH (mV)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# salinity in psu 
exo_salinity <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_SAL_PS_647937.csv"),
                               # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Salinity (PSU)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# specific conductivity in uS/cm 
exo_sp_conductivity <- read_csv(here(path, 
                                           "UAB2_B240001_YSI_EXO3_SPCOND_648943.csv"),
                                      # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Conductivity, Specific (µS/cm)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# total algae phycoerythrin in RFU 
exo_total_algae <- read_csv(here(path, 
                                       "UAB2_B240001_YSI_EXO3_TAL_PE_650659.csv"),
                                  # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Total Algae (RFU)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# total dissolved solids in mg/L 
exo_total_dissolved_solids <- read_csv(here(path,
                                          "UAB2_B240001_YSI_EXO3_TDS_MG_653051.csv"),
                                          # select columns with date, time, and the data values
                                          col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Total Dissolved Solids (mg/L)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# temperature in degrees Celcius 
exo_temp_c <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_TEMP_C_655738.csv"),
                             # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Temperature (ºC)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# temperature in degrees Fahrenheit 
exo_temp_f <- read_csv(here(path, "UAB2_B240001_YSI_EXO3_TEMP_F_659351.csv"),
                             # select columns with date, time, and the data values
                                   col_select = c("TIMESTAMP", "VALUE")) %>%
  # rename columns to match above files
  rename(datetime = TIMESTAMP,
         `Temperature (ºF)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

```

Going to move the Data Check step to once the Sonde dataframes are combined to save some time.

## Data Wrangling

Join all dataframes together so there is one 2011-2022 dataframe.

### 2011-2021 SCADA Data

Combine the data:

```{r data-wrangling-2011-2021}
# create 2011-2021 ocean data dataframe
OceanDataFrame_2011_2021 <- rbind(ocean_data_2011_2014, 
                                  ocean_data_2014,
                                  ocean_data_2014_2016,
                                  ocean_data_2016_2021)%>%
  # turn units into uS/cm instead of mS/cm
  rename(`Conductivity (µS/cm)` = `Conductivity (mS/cm)`) %>%
  mutate(`Conductivity (µS/cm)` = `Conductivity (µS/cm)`*1000)%>%
  mutate(Date = as.Date(Date)) %>%
  distinct()

```

Check the data again:

```{r check-oceandataframe-2011-2021}
OceanDataFrame_2011_2021 %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows
```

### Old Database Data: 10/21/2019 - 2/23/2022

```{r data-wrangling-old-database}
# create old database YSI EXO3 dataframe by joining the files for each variable from this instrument based on the Date and Time columns
YsiFrame <- exo_temp_f %>%
  left_join(exo_temp_c, by = join_by(Date,Time)) %>%
  left_join(exo_ph, by = join_by(Date,Time)) %>%
  left_join(exo_salinity, by = join_by(Date,Time)) %>%
  left_join(exo_conductivity, by = join_by(Date,Time)) %>%
  left_join(exo_orp, by = join_by(Date,Time)) %>%
  left_join(exo_total_algae, by = join_by(Date,Time)) %>%
  left_join(exo_do_saturated, by = join_by(Date,Time)) %>%
  left_join(exo_do, by = join_by(Date,Time)) %>%
  left_join(exo_do_local, by = join_by(Date,Time)) %>%
  left_join(exo_nlf_conductivity, by = join_by(Date,Time)) %>%
  left_join(exo_ph_mv, by = join_by(Date,Time)) %>%
  left_join(exo_sp_conductivity, by = join_by(Date,Time)) %>%
  left_join(exo_total_dissolved_solids, by = join_by(Date,Time)) %>%
  add_column(`Chlorophyll (µg/L)` = NA, .after = 9) %>%
  add_column(`Total Algae (cells/L)` = NA, .after = 10) %>%
  left_join(exo_chlorophyll, by = join_by(Date,Time)) %>%
  mutate(Date = gsub("/", "-", Date))%>%
  mutate(Date = as.Date(Date))


```

Check the Sonde data:

```{r check-sond-data}
YsiFrame %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows
```

## Combine Dataframes

Combine dataframes, keep unique rows, and move columns around so that similar variables are grouped together

```{r data-wrangling-combine-data}

OceanWaterQualityFrame <- rbind(OceanDataFrame_2011_2021, YsiFrame[, 1:12]) %>%
  left_join(YsiFrame[,c(1:2,13:19)], by = join_by(Date,Time)) %>%
  relocate(`pH (mV)`, .after = pH) %>%
  relocate(`Conductivity, Non-Linear Function (µS/cm)`, 
           `Conductivity, Specific (µS/cm)`, .after = `Conductivity (µS/cm)`) %>%
  relocate(`Dissolved Oxygen (mg/L)`, 
           `Dissolved Oxygen, Saturated (%)`, 
           `Dissolved Oxygen, Local (%)`, .after = `Conductivity, Specific (µS/cm)`) %>%
  relocate(`Chlorophyll (RFU)`, .after = `Oxygen Reduction Potential (mV)`)  %>%
  distinct()


```

Check the data one more time

```{r final-data-check}
dfCheck <- OceanWaterQualityFrame %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows
dfCheck
```

## Write Data to File

```{r write-data-to-file}
write_csv(OceanWaterQualityFrame,
          file = "Data/Clean/YSI/01_DataWrangling/01-YSI-Data-Clean-2011-2022.csv",
          col_names = TRUE)
```

Ignore:

```{r eval=FALSE, echo=FALSE, import-and-format-UTF-16-files}
# The following chunks are original code from Dan (and Renee) for EXO3 files -- can be ignored
#file path for your data directory. might work with the "Box Sync" client as well, haven't tried
path16 <- "/Users/grambihler/Documents/Data/GitHub/OceanWaterQuality/YSIExoSonde/YSI-sonde-raw-data/UTF-16-versions" 

# create ExoFrame16: use path to create list of UTF-16 files to read in, read in files in path, specify column data types, skip the first 9 rows, and specify UTF-16 encoding for files from YSI exo sonde
ExoFrame16 <- list.files(path = path16,
                       pattern = ".csv") %>%
            map_df(~read_csv(.,
                             col_types = cols(.default = col_double(),
                                              `Date (MM/DD/YYYY)` = col_date(format = '%m/%d/%Y'),
                                              `Site Name` = col_character(),
                                              `Time (HH:mm:ss)` = col_time(format = "%H:%M:%S")),
                             skip = 9,
                             locale=locale(encoding="UTF-16")))
```

```{r eval=FALSE, echo=FALSE, Set-Working-Directory}
setwd("C:/Users/fried/OneDrive/Desktop/YSI Data") #your data directory. might work with the "Box Sync" client as well, haven't tried
```

```{r eval=FALSE, echo=FALSE, File-Formatting}
files <- list.files(pattern = ".csv") #create list of files in directory

#bind files into a list. skips the first 8 lines (EXO summary data). creates a column noting the source file of each entry
exolist<-Map(cbind, lapply(files, fread,sep=",",skip=8), filename = files) 

#fread function can't work with the files as they come out of YSI (in UTF-16 format). 
#I had to re-save them in Excel to change them to UTF-8. The YSI outputs should either be converted in bulk in the future, or maybe there's a setting in the KorEXO template.

```

```{r eval=FALSE, echo=FALSE, Data-Wrangling}
#bind list entries into single merged data frame
exodata <- rbindlist(exolist,use.names=TRUE,fill=TRUE) 

#rename columns to easier names
colnames(exodata)<-c("Date","Time","Time.fracsec","Sitename","Chlor.RFU","Cond.uScm",
                     "nLF.Cond.uScm","DO.percsat","DO.perclocal","DO.mgL","ORP.mV",
                     "S.PSU","SpCond.uScm","TAL.PE.RFU","TDS.mgL","Wiperpos.V",
                     "pH","pH.mV","TempC","BattV","CablepwrV","filename") 

#I find that R does best when you break up the date column into its parts and put them back together into a date
#break date into its parts
exodata <- exodata %>%
  separate(Date,c("Month","Day","Year"),sep="/") 

#break time into its parts
exodata <- exodata %>%
  separate(Time,c("Hour","Minute","Second"))

#paste them together again and make a date
exodata$Datetime <- ymd_hms(paste(exodata$Year,exodata$Month,exodata$Day,
                                  " ",exodata$Hour,":",exodata$Minute,":",exodata$Second))

#create a date column again
exodata$Date <- as_date(date(exodata$Datetime))

#remove any duplicate rows
exodata <- distinct(exodata)
#remove any NAs
exodata <- subset(exodata,is.na(exodata$Datetime)==FALSE) 

#make any columns that might be stuck as factors numeric
exodata <- exodata %>% 
  mutate_at(c(9:25), as.numeric) 

#create a time tibble
exotbl <- as_tbl_time(exodata,index=Datetime)

#order by day
exotblm<- exotbl %>% 
  arrange(ymd_hms(exotbl$Datetime)) 
```

```{r eval=FALSE, echo=FALSE, Data-Analysis-Syntax-Example}
#this syntax can be used to aggregate the data by all sorts of periods including week, hour, etc
exoday<-exotbl %>%
  collapse_by("day") %>%
  group_by(Datetime,Date)%>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

```

```{r eval=FALSE, echo=FALSE, Plot-Example}
#create plots for data points

ggplot(exoday,aes(Date,DO.mgL))+
ggplot(exoday,aes(Date,pH))+
  geom_point(cex=1,alpha=0.5)+
  ylab(expression(paste("pH")))+
  xlab(element_blank())+
  theme_classic(base_size=20)+
  theme(legend.position = "bottom")


```
