---
title: "Ocean Data Quality Control"
collaborators: Renee Grambihler
project: OceanWaterQuality
institution: Biosphere 2 Ocean
output: html_notebook
---
# Data Quality Control

> Please Note:  If you have access to the file 02-YSI-Data-QCCutoff.csv, then you do not need to run this notebook in order to proceed with downstream data analyses. If you would like to see how the data was quality controlled, then continue.


## Import Libraries
```{r Import Libraries}
library(tidyverse)
library(lubridate)
# turns scientific notation off
options(scipen = 999)
```

## Import Raw Data
This uses csv file created in OceanDataWrangling.Rmd. You do not need to re-run that script to use this csv.
```{r Import Data}
# if you get an error when running line below, make sure to set R Markdown to run chunks in project directory: Tools > Global Options > R Markdown (in side panel) > "Evaluate chunks in directory:" drop down menu select Project

df <- read_csv("Data/Clean/YSI/01_DataWrangling/02-YSI-Data-Clean-All.csv") 

head(df)

```
## Plot Pre-QC Data
Make a quick plot of all variables using raw data (pre quality control) to see roughly what it looks like
```{r Visualize Data Before QC Performed}
# format data as long instead of wide so it is easier to facet in ggplot 
dfPivot <- df %>%
  pivot_longer(cols = `Temperature (ºF)`:`Total Dissolved Solids (mg/L)`,
               names_to = "variable",
               values_to = "variable_value")

# this plots all variables at once and can be hard to see on a small screen. You can change the code to look at one variable at a time by adding a filter line and commenting out the facet_wrap line: see option commented below
# getting list of column names from df, excluding "date" and "time" so we can use it to loop through
var_names <- colnames(df)
var_names <- var_names[! var_names %in% c("Date", "Time")]
plotList_preqc <- list()
for (i in 1:length(var_names)) {
  plotList_preqc[[i]] <- dfPivot %>%
    filter(variable == var_names[i]) %>%
    ggplot(aes(x = Date,
               y = variable_value)) +
    geom_point() +
    ylab(var_names[i])
  
  print(plotList_preqc[[i]])
}

```

## Quality Control Data
### Pre-QC Cutoffs
Implement cutoffs based on the context of our dataset that will account for sensor drift and ageing, espeically with such a large quantity of data:
- salinity <= 15 psu
- conductivity <= 25000
- pH < 5
```{r pre-qc-cutoffs}
# implement calibration cutoff values
df <- df %>%
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` < 25000, NA, `Conductivity (µS/cm)`)) %>%
  mutate(pH = ifelse(pH < 5.0, NA, pH)) %>%
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` <= 15, NA, `Salinity (PSU)`)) 
```


### Check Data Size
Can't run groups of 5000 or more observations through the QC function, so we need to check the data size. Want resulting dataframe from this chunk to have 0 observations in it. 
```{r qc-data-check}
# re-run dfPivot code to update from changes to df
dfPivot <- df %>%
  pivot_longer(cols = `Temperature (ºF)`:`Total Dissolved Solids (mg/L)`,
               names_to = "variable",
               values_to = "variable_value")

dfCheck <- dfPivot %>%
  # get the date as the month and year by taking off the last 4 digits (have to add an extra in there so it ends up being -4)
  mutate(Date = str_sub(Date, end=-4)) %>%
  unique() %>%
  na.omit() %>%
  group_by(Date, variable) %>%
  summarise(n = n()) %>%
  filter(n >= 5000)

```

### QC Function
The following code defines a function to identify data points outside input standard deviations from the montlhy mean or median for each variable and replaces the identified outliers with NA values.
At some point, this will be localized within a .R file called QCFunction.R, but for now, run the below chunk to quality control the data
```{r QC Function}
# make empty dataframe to store output from loop
# creating with just 2 columns for now since "date" and "time" columns will not change
dfQC <- data.frame(matrix(ncol = 2, nrow = nrow(df)))
colnames(dfQC) <- c("Date", "Time")
# filling in date and time columns with same values from df
dfQC <- dfQC %>%
  mutate(Date = df$Date,
         Time = df$Time)
# getting list of column names from df, excluding "date" and "time" so we can use it to loop through
var_names <- colnames(df)
var_names <- var_names[! var_names %in% c("Date", "Time")]

# loopin time
for (v in var_names) {
  # select column from df to use in calculations (represented by v in the loop) and store as var
  var <- df[[v]]
  # frame with just date, time, and var columns - all from df
  dfVar <- df[, c("Date", "Time")]
  dfVar <- cbind(dfVar, var)
  
  dfVar <- dfVar %>%
    # mutate date column so it's just year and month - this will allow us to do summary statistics by month
    # mutate(Date = paste(year(Date), week(Date), sep = "-"),
    #        stat = NA) %>%
    mutate(stat = NA) %>%
    group_by(Date) %>%
    # make columns for monthly standard deviation, mean, and median absolute deviation (mad) 
    mutate(sd_monthly = sd(as.numeric(var), na.rm = TRUE),
           mean_monthly = mean(as.numeric(var), na.rm = TRUE),
           mad_monthly = mad(as.numeric(var), na.rm = TRUE))
  
  # create vector of unique dates from dfVar. Will give just the unique year/month combos
  unique_dates <- unique(dfVar$Date)

  # test to see if the data for each month are normal
  # if normal: use mean
  # if not normal: use median
  for (d in unique_dates) {
    # filter dfVar by month and use only unique values
    dfVarNormal <- dfVar %>%
      filter(Date == d) %>%
      unique() %>%
      na.omit()
    
    # run shapiro test on var values if there are more than 3 unique values. 
    # p-value will be saved in shapiro_value if above is true, otherwise, sample will be considered non-normal, and shapiro_value will be saved as 0 
    shapiro_value <- NA
    if (length(unique(dfVarNormal$var)) <= 3) {
      shapiro_value <- 0
    } else {
      shapiro_value <- shapiro.test(as.numeric(dfVarNormal$var))$p.value
    }
    
    # if shapiro_value is not significant (data is normal), then make stat values where date == d be equal to "mean", otherwise, set them equal to "median"
    dfVar <- dfVar %>%
      mutate(stat = ifelse(Date== d & shapiro_value >= 0.05,
                           "mean",
                           "median"))
  }
  
  # calculate z-scores for each data point based on the monthly mean or median
  # if the value of stat is "mean", then a normal z-score will be calculated using the mean
  # if the value of stat is "median", then a modified z-score will be calculated using the median
  # if the denominator in the suggested calculation is going to be 0 (sd_monthly for normal and mad_monthly for modified), then either the normal z-score will be calculated, or 0 will be put in for the zscore. Some reasoning is included below: 
  dfVar <- dfVar %>%
    # create column called zscore and populate with NA values
    mutate(zscore = NA,
           # if stat is median and mad_monthly is > 0, then calculate the modified z-score for zscore column, otherwise, keep original zscore value (NA in this case). This is to avoid dividing by 0
           zscore = ifelse(stat == "median" & mad_monthly > 0,
                           0.6745 * (var - mean_monthly) / mad_monthly,
                           zscore),
           # if stat is mean and sd_monthly is > 0, then calculate normal z-score for zscore column, otherwise, keep original zscore value. This is to avoid dividing by 0.
           zscore = ifelse(stat == "mean" & sd_monthly > 0,
                           (var - mean_monthly)/sd_monthly,
                           zscore),
           # if stat is mean and sd_monthly is < or = to 0, make zscore value 0, otherwise, keep original zscore value. If the data are normal, and the standard deviation is <=0, then the z-score should be 0 anyways. This is again to avoid dividing by 0.
           zscore = ifelse(stat == "mean" & sd_monthly <= 0,
                           0,
                           zscore), 
           # if stat is median and mad_monthly is <= 0 and sd_monthly > 0, then calculate a normal z-score for zscore column, otherwise, keep original zscore value. This is to avoid dividing by 0: if median is suggested (non normal data) but the var value is not different from the median, this will give 0 as mad value (mad = constant * |var - median|; see ?mad for more details)
           zscore = ifelse(stat == "median" & mad_monthly <= 0 & sd_monthly > 0,
                           (var - mean_monthly)/sd_monthly,
                           zscore), 
           # if stat is median and mad_monthly is <= 0 and sd_monthly <= 0, zscore is 0, otherwise, keep original zscore value
           zscore = ifelse(stat == "median" & mad_monthly <= 0 & sd_monthly <= 0, 
                           0,
                           zscore))

  # if absolute value of z-score is greater than 2.00, then turn var value into NA; otherwise, leave as is
  dfVar <- dfVar %>%
    mutate(var = ifelse(abs(zscore) > 2.00,
                        NA,
                        var))
  # replace var variable with new QC'd var column from dfVar
  var <- dfVar$var
  
  # stick new QC'd var column into dfQC
  dfQC <- dfQC %>%
    cbind(var)
  
  # set the column name of dfQC that is currently called "var" to be what v currently represents in the loop. Before this, the column names for dfQC are date, time, var. After this, the column names for dfQC are date, time, temp_f, temp_c, etc. Otherwise, you'd just keep replacing the data in var with new QC'd data but not saving it to a column with the name of the variable that data represents (like pH or temp_f).
  colnames(dfQC)[colnames(dfQC) == "var"] = v
  }

```

### Save QC Data
Save QC'd data from above chunk to csv called 02-OceanWaterQualityData-QC. The next steps take the QC process further by implementing cutoff values. 
```{r Save QC Data}
write_csv(dfQC, 
          file = "Data/Clean/YSI/02_QualityControl/01-YSI-Data-QC.csv", 
          col_names = TRUE)
```


### Visualize QC Data Before Cutoffs Implemented
```{r Post-QC, Pre-Cutoffs Visualization}
# dfQC should still be in your environment if you did all of the above chunks already. If you are starting from this step, uncomment the below line: 
# dfQC <- read_csv("Data/Clean/YSI/02_QualityControl/01-YSI-Data-QC.csv")

# pivot dfQC to long format from wide so that facet wrapping is easier
dfQCPivot <- dfQC %>%
  pivot_longer(cols = `Temperature (ºF)`:`Total Dissolved Solids (mg/L)`,
               names_to = "variable",
               values_to = "variable_value")%>%
  mutate(Date = as.Date(Date, format = "%Y/%m/%d"))


plotList_postqc <- list()
for (i in 1:length(var_names)) {
  plotList_postqc[[i]] <- dfQCPivot %>%
    filter(variable == var_names[i]) %>%
    ggplot(aes(x = Date,
               y = variable_value)) +
    geom_point() +
    ylab(var_names[i]) +
    scale_x_date(date_breaks = "1 year",
                 date_labels = "%Y")
  
  print(plotList_postqc[[i]])
}

# dfQCPivot %>% 
#   filter(Date >= "2020-06-01" & Date <= "2021-01-01") %>%
#   filter(variable == "Salinity (PSU)") %>%
#   ggplot(aes(x = Date,
#                y = variable_value)) +
#     geom_point() +
#   #scale_x_datetime(date_breaks = "12 hours") +
#   scale_x_date(date_breaks = "1 month",
#                  date_labels = "%B")+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#     ylab("Temperature (ºC)")

```


### Implement Cutoffs
Based on the above plot, the variables pH, salinity_psu, and temp_c still have some remaining outliers. The below chunk will implement cutoffs to replace those outliers with NAs. The rules being used for cutoffs are:
- pH < 7.5
- salinity_psu < 21
- salinity_psu < 31 & date > "2020/01/01"
- salinity_psu < 25 & date < "2013/01/01"
- temp_c < 20 or > 28
```{r Implement Cutoffs}
# read in QCOceanWaterQualityData.csv that was created in a previous chunk if it is not already in your environment
dfQC <- read_csv("Data/Clean/YSI/02_QualityControl/01-YSI-Data-QC.csv")

# implement post-QC cutoff values
dfQCCutoff <- dfQC %>%
  mutate(`Conductivity (µS/cm)` = ifelse(`Conductivity (µS/cm)` < 10000, NA, `Conductivity (µS/cm)`)) %>%
  mutate(`Conductivity (µS/cm)` = ifelse(Date >= "2020/01/01" & Date <= "2022-01-01" & `Conductivity (µS/cm)` < 43000, NA, `Conductivity (µS/cm)`)) %>%
  mutate(pH = ifelse(pH < 7.5, NA, pH)) %>%
  mutate(`Salinity (PSU)` = ifelse(`Salinity (PSU)` < 21, NA, `Salinity (PSU)`)) %>%
  mutate(`Salinity (PSU)` = ifelse(Date >= "2020/01/01" & Date <= "2022-01-01" & `Salinity (PSU)` < 31, NA, `Salinity (PSU)`)) %>%
  mutate(`Salinity (PSU)` = ifelse(Date <= "2013/01/01" & Date >= "2012-01-01" & `Salinity (PSU)` < 25, NA, `Salinity (PSU)`)) %>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` < 20, NA, `Temperature (ºC)`))%>%
  mutate(`Temperature (ºC)` = ifelse(`Temperature (ºC)` > 28, NA, `Temperature (ºC)`))

```


## Visualize Post-QC Data
Visualize final QC'd data to see if there are anymore changes to make.
```{r Visualize Post-QC, Post-Cutoff Data}
# pivot dfQCCutoff to long format from wide so that facet wrapping is easier
dfQCCutoffPivot <- dfQCCutoff %>%
  pivot_longer(cols = `Temperature (ºF)`:`Chlorophyll (RFU)`,
               names_to = "variable",
               values_to = "variable_value")%>%
  mutate(Date = as.Date(Date, format = "%Y/%m/%d"))

# again, this plots all variables at once and can be hard to see on a small screen. You can change the code to look at one variable at a time by adding a filter line and commenting out the facet_wrap line: see option commented below
plotList_postqccutoff <- list()
for (i in 1:length(var_names)) {
  plotList_postqccutoff[[i]] <- dfQCCutoffPivot %>%
    filter(variable == var_names[i]) %>%
    ggplot(aes(x = Date,
               y = variable_value)) +
    geom_point() +
    ylab(var_names[i]) +
    scale_x_date(date_breaks = "1 year",
                 date_labels = "%Y")
  
  print(plotList_postqccutoff[[i]])
}


```

# Write Post-QC CSV
Everything looks good, so we can write the final QC'd data to a csv file.
```{r Write Post-QC csv}
write_csv(dfQCCutoff, 
          file = "Data/Clean/YSI/02_QualityControl/02-YSI-Data-QCCutoff.csv", 
          col_names = TRUE)  

```
