---
title: "Hach and YSI Figures"
collaborators: Renee Grambihler
project: "The Biosphere 2 Ocean mesocosm: (re)engineered for state-of-the-art coral reef resilience experiments"
institution: Biosphere 2 Reef Lab
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
# Hach and YSI Figures
## Input and Output Files
### Input Files
OutputData/: 
  `HachYSI-Data-Medians-All.csv`: all median time interval data combined for all colorimetry and titration nutrient data (hach) and multiparameter sonde data (ysi) for the Biosphere 2 Ocean from 2011-2024  
    - variables:   
      `Time Interval`: character; time interval at which the median is being taken, "daily", "weekly", "monthly", and "yearly" for this file  
      `Date`: date; YYYY-mm-dd  
      `Temperature (ºF)`: numeric; temperature in degrees Fahrenheit  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `pH (mV)`: numeric; pH in miliVolts  
      `Salinity (PSU)`: numeric; salinity in Potenial Salinity Units  
      `Conductivity (µS/cm)`: numeric; electrical conductivity in µSiemens/cm  
      `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity as a non-linear function in µSiemens/cm  
      `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity in µSiemens/cm  
      `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen in mg/L  
      `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen in %  
      `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen in %  
      `Oxygen Reduction Potential (mV)`: numeric; oxidative-reduction potential in miliVolts  
      `Chlorophyll (RFU)`: numeric; chlorophyll in Relative Fluorescent Units  
      `Total Algae (RFU)`: numeric; total algae in Relative Fluorescent Units  
      `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids in mg/L  
      `pH (YSI)`: numeric; pH in pH units from the YSI dataset  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units   
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.3-30.0 measured in mg/L  
      `pH_Hach`: numeric; pH in pH units from the Hach dataset  
      `Alkalinity (mg/L CaCO3)`: numeric; Calcium Carbonate (CaCO3) in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; Silica (SiO2) in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; Iron in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; Ammonia (NH3) in range 0.01-0.50 measured in mg/L  

### Output Files
Figures/: 
  `HachYSI-1x3HistoryPlot.png`: Scatter plot of temperature (c), salinity, and pH from 2011-2024 to show historical context of B2O. 
  
  `HachYSI-1x3CurrentPlot.png`: Scatter plot of temperature (c), salinity, and pH from 2021-2024 to show environmental variability of B2O in recent years.
  
  `HachYSI-NutWatQualPlot-TimeSeries.png`: Line plot with shaded ranges for Temperature (ºC), Salinity (PSU), pH, Alkalinity (mg/L CaCO3), Nitrate (mg/L), Phosphate (mg/L), Silica (mg/L), Iron (mg/L), and Dissolved Oxygen (mg/L) in B2O from 2020-2024. 
  
  `HachYSI-NutWatQualPlot-Climatology.png`: Line plot showing seasonal variation of Temperature (ºC), Salinity (PSU), pH, Alkalinity (mg/L CaCO3), Nitrate (mg/L), Phosphate (mg/L), Silica (mg/L), Iron (mg/L), and Dissolved Oxygen (mg/L) over the course of each year (january - december) from 2021-2024. 
  

```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
options(scipen = 999) 
```

# Import Data
```{r import-data}
# time series data
dfTS <- read_csv("OutputData/HachYSI-Data-Medians-All.csv") %>%
  select(-pH_Hach) %>% #take out pH data from Hach
  rename(pH = `pH (YSI)`, #rename pH data from YSI to just pH
         `Nitrate (mg/L)` = `Nitrate, Mid-Range (mg/L)`) %>% # rename Nitrate MR to just Nitrate
  # implement another cutoff value -- heat exchanger very reliable and kept temp lower than or equal to 25.5 during 2021
  mutate(`Temperature (ºC)` = ifelse(Date >= "2021-01-01" & Date < "2022-01-02" & `Temperature (ºC)` > 25.5, 
                                     NA, 
                                     `Temperature (ºC)`))
```

# B2O History 2011-2024
Scatter plot nutrient and water quality data from 2011-2024 to show historical context of B2O. 
```{r history}
# history plot 
# 1x3: Temp C, Salinity (PSU), pH; 2011-2024
plotHistory <- dfTS %>%
  # use Daily time interval
  filter(`Time Interval` == "Daily") %>%
  # keep just date, temperature celcius, salinity psu, and pH columns
  select(Date, `Temperature (ºC)`, `Salinity (PSU)`, pH) %>%
  pivot_longer(cols = `Temperature (ºC)`:pH,
               names_to = "var_name",
               values_to = "var_value") %>%
  mutate(across(var_name, ~factor(., levels=c("Temperature (ºC)","Salinity (PSU)","pH")))) %>%
  ggplot(aes(x = Date,
             y = var_value)) +
  facet_wrap(~var_name,
             ncol = 1,
             scales = "free_y",
             strip.position = "left", 
             labeller = as_labeller(c(`Temperature (ºC)` = "Temperature (ºC)", 
                                      `Salinity (PSU)` = "Salinity (PSU)",
                                      pH = "pH"))) +
  geom_point(size = 0.25) +
  scale_x_date(date_breaks = "year",
               date_labels = "%Y") +
  ylab(NULL) +
  xlab(NULL) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"),
        strip.background = element_blank(), 
        strip.placement = "outside")
  
plotHistory

# Save Plot
ggsave(filename = "HachYSI-1x3HistoryPlot.png",
       plot = plotHistory,
       path = "Figures")

```

# Environmental Variability 2021-2024
Scatter plot of temperature, salinity, and pH from 2021-2024 to show environmental variability of B2O in recent years.
```{r current}
# environmental variability plot 
# 1x3: Temp C, Salinity (PSU), pH; 2021-2024
plotCurrent <- dfTS %>%
  # use Daily time interval
  filter(`Time Interval` == "Daily") %>%
  filter(Date >= "2021-01-01") %>%
  # keep just date, temperature celcius, salinity psu, and pH columns
  select(Date, `Temperature (ºC)`, `Salinity (PSU)`, pH) %>%
  pivot_longer(cols = `Temperature (ºC)`:pH,
               names_to = "var_name",
               values_to = "var_value") %>%
  mutate(across(var_name, ~factor(., levels=c("Temperature (ºC)","Salinity (PSU)", "pH")))) %>%
  ggplot(aes(x = Date,
             y = var_value)) +
  facet_wrap(~var_name,
             ncol = 1,
             scales = "free_y",
             strip.position = "left", 
             labeller = as_labeller(c(`Temperature (ºC)` = "Temperature (ºC)", 
                                      `Salinity (PSU)` = "Salinity (PSU)",
                                      pH = "pH"))) +
  geom_point(size = 0.25) +
  scale_x_date(date_breaks = "year",
               date_labels = "%Y") +
  ylab(NULL) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"),
        strip.background = element_blank(), 
        strip.placement = "outside")
  
plotCurrent

# Save Plot
ggsave(filename = "HachYSI-1x3CurrentPlot.png",
       plot = plotCurrent,
       path = "Figures",
       width = 6,
       height = 4)

```

# Nutrients and Water Quality Time Series 2020-2024
Line plot with ranges for nutrients and water quality of B2O from 2020-2024. 
```{r nutrients-water-qual-timeseries}
# nutrients and water quality time series plot 
# Temp, Salinity, pH, Alaklinity Nitrate, Phosphate, Silica, Iron, DO; 2021-2024 
dfNutWatQual_reg <- dfTS %>%
  # use Daily time interval
  filter(`Time Interval` == "Daily") %>%
  filter(Date >= "2021-01-01")%>%
  filter(`Dissolved Oxygen (mg/L)` < 7.0) %>%
  # keep just date, alkalinity, nitrate mr, iron, and silica columns
  select(Date, `Temperature (ºC)`, `Salinity (PSU)`, pH, `Alkalinity (mg/L CaCO3)`, `Nitrate (mg/L)`, `Phosphate (mg/L)`, `Silica (mg/L)`, `Iron (mg/L)`, `Dissolved Oxygen (mg/L)`) %>%
  pivot_longer(cols = `Temperature (ºC)`:`Dissolved Oxygen (mg/L)`,
               names_to = "var_name",
               values_to = "var_value") %>%
  mutate(across(var_name, ~factor(., levels=c("Temperature (ºC)", "Salinity (PSU)", "pH", "Alkalinity (mg/L CaCO3)", "Nitrate (mg/L)", "Phosphate (mg/L)", "Silica (mg/L)", "Iron (mg/L)", "Dissolved Oxygen (mg/L)"))))

dfNutWatQual_naomit <- dfTS %>%
  # use Daily time interval
  filter(`Time Interval` == "Monthly") %>%
  filter(Date >= "2021-01-01") %>%
  filter(`Dissolved Oxygen (mg/L)` < 7.0) %>%
  # keep just date, alkalinity, nitrate mr, iron, and silica columns
  select(Date, `Temperature (ºC)`, `Salinity (PSU)`, pH, `Alkalinity (mg/L CaCO3)`, `Nitrate (mg/L)`, `Phosphate (mg/L)`, `Silica (mg/L)`, `Iron (mg/L)`, `Dissolved Oxygen (mg/L)`) %>%
  pivot_longer(cols = `Temperature (ºC)`:`Dissolved Oxygen (mg/L)`,
               names_to = "var_name",
               values_to = "var_value") %>%
  na.omit() %>%
  mutate(across(var_name, ~factor(., levels=c("Temperature (ºC)", "Salinity (PSU)", "pH", "Alkalinity (mg/L CaCO3)", "Nitrate (mg/L)", "Phosphate (mg/L)", "Silica (mg/L)", "Iron (mg/L)", "Dissolved Oxygen (mg/L)"))))

plotNutWatQual_ts <- 
    ggplot(data = dfNutWatQual_reg,
           aes(x = Date,
               y = var_value)) +
    facet_wrap(~var_name,
               ncol = 2,
               scales = "free_y",
               strip.position = "left", 
               labeller = as_labeller(c(`Temperature (ºC)` = "Temperature (ºC)",
                                        `Salinity (PSU)` = "Salinity (PSU)",
                                        pH = "pH", 
                                        `Alkalinity (mg/L CaCO3)` = "Alkalinity (mg/L CaCO3)",
                                        `Nitrate (mg/L)` = "Nitrate (mg/L)",
                                        `Phosphate (mg/L)` = "Phosphate (mg/L)", 
                                        `Silica (mg/L)` = "Silica (mg/L)",
                                        `Iron (mg/L)` = "Iron (mg/L)",
                                        `Dissolved Oxygen (mg/L)` = "Dissolved Oxygen (mg/L)"))) +
    geom_point(color = "lightgray",
               alpha = 0.4) +
    geom_line(data = dfNutWatQual_naomit,
              aes(x = Date,
               y = var_value)) +
    scale_x_date(date_breaks = "year",
                 date_labels = "%Y") +
    ylab(NULL) +
    xlab(NULL) +
    theme_bw() + 
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(colour = "black"),
          strip.background = element_blank(), 
          strip.placement = "outside") 

plotNutWatQual_ts

# Save Plot
ggsave(filename = "HachYSI-NutWatQualPlot-TimeSeries.png",
       plot = plotNutWatQual_ts,
       path = "Figures",
       height = 8)


```

# Seasonal Variation 2021-2024
Line plot showing variation over the course of a year from 2021-2024. 
```{r nutrients-water-qual-climatology}
# nutrients and water quality time series plot 
# Temp, Salinity, pH, Alaklinity Nitrate, Phosphate, Silica, Iron, DO; 2021-2024 
dfNutWatQual_s <- dfTS %>%
  # use Daily time interval
  filter(`Time Interval` == "Monthly") %>%
  filter(Date >= "2021-01-01") %>%
  filter(`Dissolved Oxygen (mg/L)` < 7.0) %>%
  # keep just columns that will be plotted
  select(Date, `Temperature (ºC)`, `Salinity (PSU)`, pH, `Alkalinity (mg/L CaCO3)`, `Nitrate (mg/L)`, `Phosphate (mg/L)`, `Silica (mg/L)`, `Iron (mg/L)`, `Dissolved Oxygen (mg/L)`) %>%
  # pivot dataframe longer bc ggplot likes that better
  pivot_longer(cols = `Temperature (ºC)`:`Dissolved Oxygen (mg/L)`,
               names_to = "var_name",
               values_to = "var_value") %>%
  # add month number (jan = 1, feb = 2, etc)
  mutate(month = month(Date, label = FALSE)) %>%
  group_by(month, var_name) %>%
  # use reframe instead of summarise bc it'll return more cols and rows 
  reframe(min = min(var_value, na.rm = TRUE),
          max = max(var_value, na.rm = TRUE),
          mean = mean(var_value, na.rm = TRUE),
          q1 = quantile(var_value, 0.25, na.rm = TRUE),
          q2 = quantile(var_value, 0.75, na.rm = TRUE)) %>%
  mutate(Date = as.Date(paste(2021, month, 01, sep = "-"))) %>%
  mutate(across(var_name, ~factor(., levels=c("Temperature (ºC)", "Salinity (PSU)", "pH", "Alkalinity (mg/L CaCO3)", "Nitrate (mg/L)", "Phosphate (mg/L)", "Silica (mg/L)", "Iron (mg/L)", "Dissolved Oxygen (mg/L)")))) 

# plot 
plotNutWatQual_s <- dfNutWatQual_s %>%
  ggplot(aes(x = as.Date(yday(Date), "2021-01-01"),
             y = mean)) +
  facet_wrap(~var_name,
             ncol = 2,
             scales = "free_y",
             strip.position = "left", 
             labeller = as_labeller(c(`Temperature (ºC)` = "Temperature (ºC)",
                                      `Salinity (PSU)` = "Salinity (PSU)",
                                      pH = "pH", 
                                      `Alkalinity (mg/L CaCO3)` = "Alkalinity (mg/L CaCO3)",
                                      `Nitrate (mg/L)` = "Nitrate (mg/L)",
                                      `Phosphate (mg/L)` = "Phosphate (mg/L)", 
                                      `Silica (mg/L)` = "Silica (mg/L)",
                                      `Iron (mg/L)` = "Iron (mg/L)",
                                      `Dissolved Oxygen (mg/L)` = "Dissolved Oxygen (mg/L)"))) +
  geom_line() +
  geom_ribbon(aes(ymin = q1,
                  ymax = q2), 
              color = "lightgray",
              alpha = 0.3) +
  scale_x_date(date_breaks = "month",
               date_labels = "%B") +
  scale_y_continuous(breaks = waiver(), n.breaks = 6) +
  ylab(NULL) +
  xlab(NULL) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(), 
        strip.placement = "outside") 
  
plotNutWatQual_s

# Save Plot
ggsave(filename = "HachYSI-NutWatQualPlot-Climatology.png",
       plot = plotNutWatQual_s,
       path = "Figures",
       height = 8)


```
