---
title: "Ocean Data Quality Control"
collaborators: Renee Grambihler
project: OceanWaterQuality
institution: Biosphere 2 Ocean
output: html_notebook
---
# Data Quality Control

> Please Note:  If you have access to the file QCOceanWaterQualityFrame.csv, then you do not need to run this notebook in order to proceed with downstream data analyses. If you would like to see how the data was quality controlled, then continue.


## Import Libraries
```{r Import Libraries}
library(tidyverse)
library(lubridate)
library(here)
# source("./QCFunction.R")

# turns scientific notation off
options(scipen = 999)
```

## Import Raw Data
```{r Import Data}
# uses csv file created in OceanDataWrangling.Rmd. do not need to re-run that file to use this csv
# if you get an error when running line below, make sure to set R Markdown to run chunks in project directory: Tools > Global Options > R Markdown (in side panel) > "Evaluate chunks in directory:" drop down menu select Project

df <- read_csv("Data/CleanData/CleanData_YSI/01_DataWrangling/OceanWaterQualityFrame.csv")

```
## Plot Pre-QC Data
```{r Visualize Data Before QC Performed}
# make quick plot of all variables using raw data (pre quality control) to see roughly what it looks like

# format data as long instead of wide so it is easier to facet in ggplot 
dfPivot <- df %>%
  pivot_longer(cols = temp_f:total_dissolved_solids_mgL,
               names_to = "variable",
               values_to = "variable_value")%>%
  mutate(date = as.Date(date, format = "%Y/%m/%d"))

# this shows all variables at once and can be hard to see on a small screen. You can change the code to look at one variable at a time by adding a filter line and commenting out the facet_wrap line: see option commented below
AllVariablePlotPreQC <- dfPivot %>%
  #filter(variable == "pH") %>%
  ggplot(aes(x = date, 
             y = variable_value)) +
  geom_line() +
  # can comment out below line if filtering by one variable
  facet_wrap(~variable, ncol = 3, scales = "free")+
  scale_x_date(breaks = waiver(), date_breaks = "1 year",
               date_labels = "%b %Y") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = margin(0.25, 0.45, 0.25, 0.25, "inch"))+
  theme(strip.background = element_rect(fill = "white"))+
  labs(title = "B2O Water Quality by Variable from 2011-2023",
       subtitle = "Pre-QC: All Outliers Still Present")+
  xlab("Date")+
  ylab("Nutrient Value")

AllVariablePlotPreQC

```

## Quality Control Data
### QC Function
```{r QC Function}
# The following code defines a function to identify data points outside input standard deviations from the mean or median for each nutrient at each location and replaces the identified outliers with NA values
# At some point, this will be localized within a .R file called QCFunction.R, but for now, run the below chunk to quality control the data

# make empty dataframe to store QC'd data in 
# with same column names and number of rows as df
dfQC <- data.frame(matrix(ncol = 2, nrow = nrow(df)))
colnames(dfQC) <- c("date", "time")
# filling in date and time columns with same values from df
dfQC <- dfQC %>%
  mutate(date = df$date,
         time = df$time)
# getting list of column names, excluding "date" and "time" 
var_names <- colnames(df)
var_names <- var_names[! var_names %in% c("date", "time")]

# loopin time
for (v in var_names) {
  # create frame with just date, time, and var columns
  var <- df[[v]]
  
  dfVar <- df[, c("date", "time")]
  dfVar <- cbind(dfVar, var)
  
  dfVar <- dfVar %>%
    mutate(date = format(as.Date(date), "%Y-%m"),#mutate date column so it's just year and month
           stat = NA) %>%
    group_by(date) %>%
    mutate(sd_monthly = sd(as.numeric(var), na.rm = TRUE),
           mean_monthly = mean(as.numeric(var), na.rm = TRUE),
           mad_monthly = mad(as.numeric(var), na.rm = TRUE))
  
  # create vector of unique dates from dfVar. Will give just the unique year/month combos
  unique_dates <- unique(dfVar$date)

  for (d in unique_dates) {
    # filter dfVar by month and use only unique values
    dfVarNormal <- dfVar %>%
      filter(date == d) %>%
      unique()
    
    # run shapiro test on var values if there are more than 3 unique values. 
    # p-value will be saved in shapiro_value if above is true, otherwise, sample will be considered non-normal, and shapiro_value will be saved as 0 
    shapiro_value <- NA
    if (length(unique(dfVarNormal$var)) <= 3) {
      shapiro_value <- 0
    } else {
      shapiro_value <- shapiro.test(as.numeric(dfVarNormal$var))$p.value
    }
    
    # if shapiro_value is not significant, then make stat values where date == d be equal to "mean", otherwise, set them equal to "median"
    dfVar <- dfVar %>%
      mutate(stat = ifelse(date == d & shapiro_value >= 0.05,
                           "mean",
                           "median"))
  }
  
  dfVar <- dfVar %>%
    mutate(zscore = NA,
           zscore = ifelse(stat == "median" & mad_monthly > 0,
                           0.6745 * (var - mean_monthly) / mad_monthly,
                           zscore),
           zscore = ifelse(stat == "mean" & sd_monthly > 0,
                           (var - mean_monthly)/sd_monthly,
                           zscore),
           zscore = ifelse(stat == "mean" & sd_monthly <= 0,
                           0,
                           zscore), 
           zscore = ifelse(stat == "median" & mad_monthly <= 0 & sd_monthly > 0,
                           (var - mean_monthly)/sd_monthly,
                           zscore), 
           zscore = ifelse(stat == "median" & mad_monthly <= 0 & sd_monthly <= 0, 
                           0,
                           zscore))

  # if absolute value of z-score is greater than 2.00, then turn var value into NA; otherwise, leave as is
  dfVar <- dfVar %>%
    mutate(var = ifelse(abs(zscore) > 2.00,
                        NA,
                        var))
  
  var <- dfVar$var
  
  dfQC <- dfQC %>%
    cbind(var)
  
  colnames(dfQC)[colnames(dfQC) == "var"] = v
  }

```

### Save QC Data
```{r Save QC Data}
write_csv(dfQC, 
          file = "Data/CleanData/CleanDataYSI/02_QualityControl/QCOceanWaterQualityFrame.csv", 
          col_names = TRUE)
```


### Visualize QC Data Before Cutoffs Implemented
```{r Post-QC, Pre-Cutoffs Visualization}
OceanWaterQualityFramePivot <- dfQC %>%
  pivot_longer(cols = temp_f:total_dissolved_solids_mgL,
               names_to = "variable",
               values_to = "variable_value")%>%
  mutate(date = as.Date(date, format = "%Y/%m/%d"))

AllVariablePlotPostQC <- OceanWaterQualityFramePivot %>%
  ggplot(aes(x = date, 
             y = variable_value)) +
  geom_line() +
  facet_wrap(~variable, ncol = 3, scales = "free")+
  scale_x_date(breaks = waiver(), date_breaks = "1 year",
               date_labels = "%b %Y") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = margin(0.25, 0.45, 0.25, 0.25, "inch"))+
  theme(strip.background = element_rect(fill = "white"))+
  labs(title = "B2O Water Quality by Variable from 2011-2023",
       subtitle = "Post-QC: Removed Values 2 Standard Deviations Outside Mean")+
  xlab("Date")+
  ylab("Nutrient Value")

AllVariablePlotPostQC

```


### Implement Cutoffs
```{r Implement Cutoffs}
# read in QCOceanWaterQualityFrame.csv that was created in a previous chunk
QCOceanWaterQualityFrame <- read_csv("Data/CleanData/CleanData_YSI/02_QualityControl/QCOceanWaterQualityFrame.csv")

# implement calibration cutoff values
CutQCOceanWaterQualityFrame <- QCOceanWaterQualityFrame %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y")) %>%
  mutate(pH = ifelse(pH < 7.5, NA, pH)) %>%
  mutate(salinity_psu = ifelse(salinity_psu < 21, NA, salinity_psu)) %>%
  mutate(salinity_psu = ifelse(date > "2020/01/01" & salinity_psu < 31, NA, salinity_psu)) %>%
  mutate(salinity_psu = ifelse(date < "2013/01/01" & salinity_psu < 24, NA, salinity_psu)) %>%
  mutate(temp_c = ifelse(temp_c < 20 | temp_c > 28, NA, temp_c))

```


## Visualize Post-QC Data
```{r Visualize Post-QC, Post-Cutoff Data}
CutQCOceanWaterQualityFramePivot <- CutQCOceanWaterQualityFrame %>%
  pivot_longer(cols = temp_f:total_dissolved_solids_mgL,
               names_to = "variable",
               values_to = "variable_value")%>%
  mutate(date = as.Date(date, format = "%Y/%m/%d"))

AllVariablePlotPostQCCutoff <- CutQCOceanWaterQualityFramePivot %>%
  ggplot(aes(x = date, 
             y = variable_value)) +
  geom_line() +
  facet_wrap(~variable, ncol = 3, scales = "free")+
  scale_x_date(breaks = waiver(), date_breaks = "1 year",
               date_labels = "%b %Y") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.margin = margin(0.25, 0.45, 0.25, 0.25, "inch"))+
  theme(strip.background = element_rect(fill = "white"))+
  labs(title = "B2O Water Quality by Variable from 2011-2023",
       subtitle = "Post-QC: Removed Values 2 Standard Deviations Outside Mean and Cutoff Values")+
  xlab("Date")+
  ylab("Nutrient Value")

AllVariablePlotPostQCCutoff

```

# Write Post-QC CSV
```{r Write Post-QC csv}
write_csv(CutQCOceanWaterQualityFrame, 
          file = "Data/CleanData/CleanData_YSI/02_QualityControl/QCOceanWaterQualityFrame.csv", 
          col_names = TRUE)  

```

