---
title: "YSI Data Descriptive Statistics"
collaborators: Renee Grambihler
project: OceanWaterQuality
institution: Biosphere 2 Ocean
output: html_notebook
---
# Descriptive Statistics
## Import Libraries
```{r Import Libraries}
library(tidyverse)
library(tibbletime) 
library(lubridate)
options(scipen = 999)
```

## Import Data
Read in fully QC'd data
```{r Import Data}
df <- read_csv("Data/CleanData/CleanData_YSI/02_QualityControl/03-OceanWaterQualityData-QCCutoff.csv")

```

## Minor Data Manipulation
Add datetime column and create timetibble ordered by day
```{r data manip}
# combine date and time into datetime column 
dfEdit <- df %>%
  mutate(datetime = paste(date, time), .before = date,
         datetime = as.POSIXct(datetime)) %>% #make sure datetime is POSIXct class
  distinct() %>% #remove duplicate rows from df
  subset(is.na(datetime)==FALSE) #remove NAs from datetime
 
dfEditPivot <- dfEdit %>%
  pivot_longer(cols = temp_f:total_dissolved_solids_mgL,
               names_to = "variable",
               values_to = "variable_value")

# inspect df to see what the classes of each column are
str(dfEditPivot)

```

## Whole Time Series Statistics
Summary statistics over entire time period of data
```{r whole time series stats}
dfWhole <- dfEditPivot %>%
  reframe(mean = mean(variable_value, na.rm = TRUE), 
          median = median(variable_value, na.rm = TRUE),
          std = sd(variable_value, na.rm = TRUE),
          IQR = IQR(variable_value, na.rm = TRUE),
          min = min(variable_value, na.rm = TRUE),
          max = max(variable_value, na.rm = TRUE),
          .by = variable)%>%
   mutate(across(where(is.numeric), round, 2))

dfWhole[sapply(dfWhole, is.infinite)] <- NA #turn infinite values into NAs that were introduced when running min and max on variables that did not have any data

write_csv(dfWhole, "Data/CleanData/CleanData_YSI/03_DescriptiveStatistics/WholeSeriesDescriptiveStats.csv")

```

## Yearly Statistics
Summary statistics by year
```{r stats by year}
dfYear <- dfEditPivot %>%
  mutate(year = year(datetime))%>%
  group_by(year, variable) %>%
  reframe(mean = mean(variable_value, na.rm = TRUE), 
          median = median(variable_value, na.rm = TRUE),
          std = sd(variable_value, na.rm = TRUE),
          IQR = IQR(variable_value, na.rm = TRUE),
          Q1 = quantile(variable_value, 0.25, na.rm = TRUE), 
          Q2 = quantile(variable_value, 0.75, na.rm = TRUE),
          min = min(variable_value, na.rm = TRUE),
          max = max(variable_value, na.rm = TRUE))%>%
  mutate(across(where(is.numeric), round, 2))

dfYear[sapply(dfYear, is.infinite)] <- NA #turn infinite values into NAs that were introduced when running min and max on variables that did not have any data

write_csv(dfYear, "Data/CleanData/CleanData_YSI/03_DescriptiveStatistics/YearlyDescriptiveStats.csv")

```

## Monthly Statistics
Summary statistics by month
```{r stats by month}
dfMonth <- dfEditPivot %>%
  mutate(year = year(datetime)) %>%
  mutate(month = month(datetime, label = TRUE, abbr = FALSE)) %>%
  group_by(year, month, variable) %>%
  reframe(mean = mean(variable_value, na.rm = TRUE), 
          median = median(variable_value, na.rm = TRUE),
          std = sd(variable_value, na.rm = TRUE),
          IQR = IQR(variable_value, na.rm = TRUE),
          min = min(variable_value, na.rm = TRUE),
          max = max(variable_value, na.rm = TRUE))%>%
  mutate(across(where(is.numeric), round, 2))

dfMonth[sapply(dfMonth, is.infinite)] <- NA #turn infinite values into NAs that were introduced when running min and max on variables that did not have any data

write_csv(dfMonth, "Data/CleanData/CleanData_YSI/03_DescriptiveStatistics/MonthlyDescriptiveStats.csv")

```

## Daily Statistics
Summary statistics by day
```{r stats by day}
dfDay <- dfEditPivot %>%
  mutate(year = year(datetime)) %>%
  mutate(month = month(datetime, label = TRUE, abbr = FALSE)) %>%
  mutate(day = day(datetime)) %>% #if want days of week instead of numbers of month, use wday()
  group_by(year, month, day, variable) %>%
  reframe(mean = mean(variable_value, na.rm = TRUE), 
          median = median(variable_value, na.rm = TRUE),
          std = sd(variable_value, na.rm = TRUE),
          IQR = IQR(variable_value, na.rm = TRUE),
          min = min(variable_value, na.rm = TRUE),
          max = max(variable_value, na.rm = TRUE))%>%
  mutate(across(where(is.numeric), round, 2))

dfDay[sapply(dfDay, is.infinite)] <- NA #turn infinite values into NAs that were introduced when running min and max on variables that did not have any data

write_csv(dfDay, "Data/CleanData/CleanData_YSI/03_DescriptiveStatistics/DailyDescriptiveStats.csv")

```

### Example code from Dan
```{r example code}
hoboph<- distinct(hoboph)#remove any duplicate rows
hoboph<-subset(hoboph,is.na(hoboph$Datetime)==FALSE) #remove any NAs

hoboph <- hoboph %>% 
  mutate_at(c(2:4), as.numeric) #make any columns that might be stuck as factors numeric
hoboph<-as_tbl_time(hoboph,index=Datetime) #create a time tibble
hobotbl<- hoboph %>% arrange(ymd_hms(hoboph$Datetime)) #order by day

#this syntax can be used to aggregate the data by all sorts of periods including week, hour, etc
hobophday<-hobotbl %>%
  collapse_by("day") %>%
  group_by(Datetime,Date)%>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```

