---
title: "Hach and YSI Descriptive Statistics"
collaborators: Renee Grambihler
project: "The Biosphere 2 Ocean mesocosm: (re)engineered for state-of-the-art coral reef resilience experiments"
institution: Biosphere 2 Reef Lab
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
# Hach and YSI Data Wrangling
## Input and Output Files
### Input Files
InputData/Clean/02-QualityControl: 
  `YSI-Data-QC-Cutoff.csv`: all Biosphere 2 Ocean remote sensing multiparameter sonde water quality data after being run through quality control parameters and cutoff values 
    - variables: 
            `Date`: date  
            `Time`: time  
            `Temperature (ºF)`: numeric; temperature value in degrees Fahrenheit  
            `Temperature (ºC)`: numeric; temperature value in degrees Celcius  
            `pH`: numeric; pH value in pH units  
            `pH (mV)`: numeric; pH value in miliVolts  
            `Salinity (PSU)`: numeric; salinity value in Potential Salinity Units  
            `Conductivity (µS/cm)`: numeric; electrical conductivity value in µSiemens/cm  
            `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity value as a non-linear function in µSiemens/cm  
            `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity value in µSiemens/cm  
            `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen value in mg/L  
            `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen value in %  
            `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen value in %  
            `Oxygen Reduction Potential (mV)`: numeric; oxidation-reduction potential value in miliVolts  
            `Chlorophyll (RFU)`: numeric; chlorophyll value in RFU  
            `Total Algae (RFU)`: numeric; total algae phycoerythrin value in RFU  
            `Chlorophyll (µg/L)`: numeric; chlorophyll value in µg/L  
            `Total Algae (cells/L)`: numeric; total algae phycoerythrin value in cells/L
            `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids value in mg/L  
        
  `Hach-Data-QC.csv`: Hach nutrient water quality data after being run through quality control parameters
    - variables:  
      `Date`: date in mm/dd/yy  
      `Location`: character; location of measurements; Ocean, Mangroves, Raceway 1, or Raceway 2  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
      `pH`: pH in pH units  
      `Alkalinity (mg/L CaCO3)`: numeric; calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; iron concentration in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  

### Output Files
OutputData/DescriptiveStatistics: 
  `HachYSI-Data-Medians-Daily.csv`: daily median of all colorimetry and titration nutrient data (hach) and multiparameter sonde data (ysi) for the Biosphere 2 Ocean from 2011-2024  
    - variables:   
      `Time Interval`: character; time interval at which the median is being taken, "daily" for this file  
      `Date`: date; YYYY-mm-dd  
      `Temperature (ºF)`: numeric; temperature in degrees Fahrenheit  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `pH (mV)`: numeric; pH in miliVolts  
      `Salinity (PSU)`: numeric; salinity in Potenial Salinity Units  
      `Conductivity (µS/cm)`: numeric; electrical conductivity in µSiemens/cm  
      `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity as a non-linear function in µSiemens/cm  
      `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity in µSiemens/cm  
      `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen in mg/L  
      `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen in %  
      `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen in %  
      `Oxygen Reduction Potential (mV)`: numeric; oxidative-reduction potential in miliVolts  
      `Chlorophyll (RFU)`: numeric; chlorophyll in Relative Fluorescent Units  
      `Total Algae (RFU)`: numeric; total algae in Relative Fluorescent Units  
      `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids in mg/L  
      `pH (YSI)`: numeric; pH in pH units from the YSI dataset  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units   
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.3-30.0 measured in mg/L  
      `pH_Hach`: numeric; pH in pH units from the Hach dataset  
      `Alkalinity (mg/L CaCO3)`: numeric; Calcium Carbonate (CaCO3) in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; Silica (SiO2) in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; Iron in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; Ammonia (NH3) in range 0.01-0.50 measured in mg/L  
  
  `HachYSI-Data-Medians-Monthly.csv`: monthly median of all colorimetry and titration nutrient data (hach) and multiparameter sonde data (ysi) for the Biosphere 2 Ocean from 2011-2024  
    - variables:   
      `Time Interval`: character; time interval at which the median is being taken, "monthly" for this file  
      `Date`: date; YYYY-mm-dd  
      `Temperature (ºF)`: numeric; temperature in degrees Fahrenheit  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `pH (mV)`: numeric; pH in miliVolts  
      `Salinity (PSU)`: numeric; salinity in Potenial Salinity Units  
      `Conductivity (µS/cm)`: numeric; electrical conductivity in µSiemens/cm  
      `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity as a non-linear function in µSiemens/cm  
      `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity in µSiemens/cm  
      `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen in mg/L  
      `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen in %  
      `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen in %  
      `Oxygen Reduction Potential (mV)`: numeric; oxidative-reduction potential in miliVolts  
      `Chlorophyll (RFU)`: numeric; chlorophyll in Relative Fluorescent Units  
      `Total Algae (RFU)`: numeric; total algae in Relative Fluorescent Units  
      `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids in mg/L  
      `pH (YSI)`: numeric; pH in pH units from the YSI dataset  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units   
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.3-30.0 measured in mg/L  
      `pH_Hach`: numeric; pH in pH units from the Hach dataset  
      `Alkalinity (mg/L CaCO3)`: numeric; Calcium Carbonate (CaCO3) in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; Silica (SiO2) in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; Iron in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; Ammonia (NH3) in range 0.01-0.50 measured in mg/L  
  
  `HachYSI-Data-Medians-Weekly.csv`: weekly median of all colorimetry and titration nutrient data (hach) and multiparameter sonde data (ysi) for the Biosphere 2 Ocean from 2011-2024  
    - variables:   
      `Time Interval`: character; time interval at which the median is being taken, "weekly" for this file  
      `Date`: date; YYYY-mm-dd  
      `Temperature (ºF)`: numeric; temperature in degrees Fahrenheit  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `pH (mV)`: numeric; pH in miliVolts  
      `Salinity (PSU)`: numeric; salinity in Potenial Salinity Units  
      `Conductivity (µS/cm)`: numeric; electrical conductivity in µSiemens/cm  
      `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity as a non-linear function in µSiemens/cm  
      `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity in µSiemens/cm  
      `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen in mg/L  
      `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen in %  
      `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen in %  
      `Oxygen Reduction Potential (mV)`: numeric; oxidative-reduction potential in miliVolts  
      `Chlorophyll (RFU)`: numeric; chlorophyll in Relative Fluorescent Units  
      `Total Algae (RFU)`: numeric; total algae in Relative Fluorescent Units  
      `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids in mg/L  
      `pH (YSI)`: numeric; pH in pH units from the YSI dataset  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units   
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.3-30.0 measured in mg/L  
      `pH_Hach`: numeric; pH in pH units from the Hach dataset  
      `Alkalinity (mg/L CaCO3)`: numeric; Calcium Carbonate (CaCO3) in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; Silica (SiO2) in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; Iron in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; Ammonia (NH3) in range 0.01-0.50 measured in mg/L  
  
  `HachYSI-Data-Medians-Yearly.csv`: yearly median of all colorimetry and titration nutrient data (hach) and multiparameter sonde data (ysi) for the Biosphere 2 Ocean from 2011-2024  
    - variables:   
      `Time Interval`: character; time interval at which the median is being taken, "yearly" for this file  
      `Date`: date; YYYY-mm-dd  
      `Temperature (ºF)`: numeric; temperature in degrees Fahrenheit  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `pH (mV)`: numeric; pH in miliVolts  
      `Salinity (PSU)`: numeric; salinity in Potenial Salinity Units  
      `Conductivity (µS/cm)`: numeric; electrical conductivity in µSiemens/cm  
      `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity as a non-linear function in µSiemens/cm  
      `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity in µSiemens/cm  
      `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen in mg/L  
      `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen in %  
      `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen in %  
      `Oxygen Reduction Potential (mV)`: numeric; oxidative-reduction potential in miliVolts  
      `Chlorophyll (RFU)`: numeric; chlorophyll in Relative Fluorescent Units  
      `Total Algae (RFU)`: numeric; total algae in Relative Fluorescent Units  
      `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids in mg/L  
      `pH (YSI)`: numeric; pH in pH units from the YSI dataset  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units   
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.3-30.0 measured in mg/L  
      `pH_Hach`: numeric; pH in pH units from the Hach dataset  
      `Alkalinity (mg/L CaCO3)`: numeric; Calcium Carbonate (CaCO3) in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; Silica (SiO2) in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; Iron in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; Ammonia (NH3) in range 0.01-0.50 measured in mg/L  
  
  `HachYSI-Data-Medians-All.csv`: all median time interval data combined for all colorimetry and titration nutrient data (hach) and multiparameter sonde data (ysi) for the Biosphere 2 Ocean from 2011-2024  
    - variables:   
      `Time Interval`: character; time interval at which the median is being taken, "daily", "weekly", "monthly", and "yearly" for this file  
      `Date`: date; YYYY-mm-dd  
      `Temperature (ºF)`: numeric; temperature in degrees Fahrenheit  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `pH (mV)`: numeric; pH in miliVolts  
      `Salinity (PSU)`: numeric; salinity in Potenial Salinity Units  
      `Conductivity (µS/cm)`: numeric; electrical conductivity in µSiemens/cm  
      `Conductivity, Non-Linear Function (µS/cm)`: numeric; electrical conductivity as a non-linear function in µSiemens/cm  
      `Conductivity, Specific (µS/cm)`: numeric; specific electrical conductivity in µSiemens/cm  
      `Dissolved Oxygen (mg/L)`: numeric; dissolved oxygen in mg/L  
      `Dissolved Oxygen, Saturated (%)`: numeric; saturated dissolved oxygen in %  
      `Dissolved Oxygen, Local (%)`: numeric; local dissolved oxygen in %  
      `Oxygen Reduction Potential (mV)`: numeric; oxidative-reduction potential in miliVolts  
      `Chlorophyll (RFU)`: numeric; chlorophyll in Relative Fluorescent Units  
      `Total Algae (RFU)`: numeric; total algae in Relative Fluorescent Units  
      `Total Dissolved Solids (mg/L)`: numeric; total dissolved solids in mg/L  
      `pH (YSI)`: numeric; pH in pH units from the YSI dataset  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units   
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; Nitrate (NO3-) from 0.3-30.0 measured in mg/L  
      `pH_Hach`: numeric; pH in pH units from the Hach dataset  
      `Alkalinity (mg/L CaCO3)`: numeric; Calcium Carbonate (CaCO3) in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; Silica (SiO2) in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; Iron in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; Ammonia (NH3) in range 0.01-0.50 measured in mg/L  


```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
options(scipen = 999)
```

## Import Data
```{r import data}
# use 02-YSI-Data-QCCutoff.csv
dfYSI <- read_csv("InputData/Clean/YSI/02_QualityControl/02-YSI-Data-QCCutoff.csv")

# use 02-Hach-Data-Clean-QC.csv
dfHach <- read_csv("InputData/Clean/02-QualityControl/Hach-Data-QC.csv")

```

# Data Formatting
## YSI
### Daily Medians
YSI Exo3 Sonde records data every 15 minutes, which means there are multiple data points for each day. To use this data with the Hach data, the median value for each day should be used. Hach data is taken once per day (Mondays and Fridays usually)
```{r data format: ysi daily medians}
dfYSI_daily <- dfYSI %>%
  # create year, month, week columns
  mutate(year = year(Date)) %>%
  mutate(month = month(Date)) %>%
 # mutate(week = week(Date)) %>%
  mutate(day = day(Date)) %>%
  # group by year, month, day to take comparisons of each week within the time series
  group_by(year, month, day) %>%
  # take daily medians 
  mutate(`Temperature (ºF)` = median(`Temperature (ºF)`, na.rm = TRUE)) %>%
  mutate(`Temperature (ºC)` = median(`Temperature (ºC)`, na.rm = TRUE)) %>%
  mutate(`pH (YSI)` = median(pH, na.rm = TRUE)) %>%
  mutate(`pH (mV)` = median(`pH (mV)`, na.rm = TRUE)) %>%
  mutate(`Salinity (PSU)` = median(`Salinity (PSU)`, na.rm = TRUE)) %>%
  mutate(`Conductivity (µS/cm)` = median(`Conductivity (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Non-Linear Function (µS/cm)` = median(`Conductivity, Non-Linear Function (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Specific (µS/cm)` = median(`Conductivity, Specific (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen (mg/L)` = median(`Dissolved Oxygen (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Saturated (%)` = median(`Dissolved Oxygen, Saturated (%)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Local (%)` = median(`Dissolved Oxygen, Local (%)`, na.rm = TRUE)) %>%
  mutate(`Oxygen Reduction Potential (mV)` = median(`Oxygen Reduction Potential (mV)`, na.rm = TRUE)) %>%
  mutate(`Chlorophyll (RFU)` = median(`Chlorophyll (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Algae (RFU)` = median(`Total Algae (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Dissolved Solids (mg/L)` = median(`Total Dissolved Solids (mg/L)`, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-Time, -Date, -`Chlorophyll (µg/L)`, -`Total Algae (cells/L)`, -pH) %>%
  distinct() %>%
  mutate(Date = paste(year, month, day, sep = "-"),
         Date = as.Date(Date)) %>%
  ungroup() %>%
  select(-year, -month, -day) %>%
  relocate(Date, .before = `Temperature (ºF)`) 

```

### Weekly Medians
```{r data format: ysi weekly medians}
dfYSI_weekly <- dfYSI %>%
  # create year, month, week columns
  mutate(date_week = ceiling_date(as.Date(Date), "week")-1) %>%
  # group by year, month, week
  group_by(date_week) %>%
  # take weekly medians 
  mutate(`Temperature (ºF)` = median(`Temperature (ºF)`, na.rm = TRUE)) %>%
  mutate(`Temperature (ºC)` = median(`Temperature (ºC)`, na.rm = TRUE)) %>%
  mutate(`pH (YSI)` = median(pH, na.rm = TRUE)) %>%
  mutate(`pH (mV)` = median(`pH (mV)`, na.rm = TRUE)) %>%
  mutate(`Salinity (PSU)` = median(`Salinity (PSU)`, na.rm = TRUE)) %>%
  mutate(`Conductivity (µS/cm)` = median(`Conductivity (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Non-Linear Function (µS/cm)` = median(`Conductivity, Non-Linear Function (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Specific (µS/cm)` = median(`Conductivity, Specific (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen (mg/L)` = median(`Dissolved Oxygen (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Saturated (%)` = median(`Dissolved Oxygen, Saturated (%)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Local (%)` = median(`Dissolved Oxygen, Local (%)`, na.rm = TRUE)) %>%
  mutate(`Oxygen Reduction Potential (mV)` = median(`Oxygen Reduction Potential (mV)`, na.rm = TRUE)) %>%
  mutate(`Chlorophyll (RFU)` = median(`Chlorophyll (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Algae (RFU)` = median(`Total Algae (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Dissolved Solids (mg/L)` = median(`Total Dissolved Solids (mg/L)`, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-Time, -Date, -`Chlorophyll (µg/L)`, -`Total Algae (cells/L)`, -pH) %>%
  distinct() %>%
  ungroup() %>%
  rename(Date = date_week) %>%
  relocate(Date, .before = `Temperature (ºF)`) 

```

### Monthly Medians
```{r data format: ysi monthly medians}
dfYSI_monthly <- dfYSI %>%
  # create date_month column with Date of last day of each month of each year
  mutate(date_month = ceiling_date(as.Date(Date), "month")-1) %>%
  # group by date_month
  group_by(date_month) %>%
  # take monthly medians 
  mutate(`Temperature (ºF)` = median(`Temperature (ºF)`, na.rm = TRUE)) %>%
  mutate(`Temperature (ºC)` = median(`Temperature (ºC)`, na.rm = TRUE)) %>%
  mutate(`pH (YSI)` = median(pH, na.rm = TRUE)) %>%
  mutate(`pH (mV)` = median(`pH (mV)`, na.rm = TRUE)) %>%
  mutate(`Salinity (PSU)` = median(`Salinity (PSU)`, na.rm = TRUE)) %>%
  mutate(`Conductivity (µS/cm)` = median(`Conductivity (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Non-Linear Function (µS/cm)` = median(`Conductivity, Non-Linear Function (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Specific (µS/cm)` = median(`Conductivity, Specific (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen (mg/L)` = median(`Dissolved Oxygen (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Saturated (%)` = median(`Dissolved Oxygen, Saturated (%)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Local (%)` = median(`Dissolved Oxygen, Local (%)`, na.rm = TRUE)) %>%
  mutate(`Oxygen Reduction Potential (mV)` = median(`Oxygen Reduction Potential (mV)`, na.rm = TRUE)) %>%
  mutate(`Chlorophyll (RFU)` = median(`Chlorophyll (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Algae (RFU)` = median(`Total Algae (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Dissolved Solids (mg/L)` = median(`Total Dissolved Solids (mg/L)`, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-Time, -Date, -`Chlorophyll (µg/L)`, -`Total Algae (cells/L)`, -pH) %>%
  distinct() %>%
  ungroup() %>%
  rename(Date = date_month) %>%
  relocate(Date, .before = `Temperature (ºF)`) 

```

### Yearly Medians
```{r data format: ysi yearly medians}
dfYSI_yearly <- dfYSI %>%
  # create date_year column with Date of last day of each year
  mutate(date_year = ceiling_date(as.Date(Date), "year")-1) %>%
  # group by year
  group_by(date_year) %>%
  # take yearly medians 
  mutate(`Temperature (ºF)` = median(`Temperature (ºF)`, na.rm = TRUE)) %>%
  mutate(`Temperature (ºC)` = median(`Temperature (ºC)`, na.rm = TRUE)) %>%
  mutate(`pH (YSI)` = median(pH, na.rm = TRUE)) %>%
  mutate(`pH (mV)` = median(`pH (mV)`, na.rm = TRUE)) %>%
  mutate(`Salinity (PSU)` = median(`Salinity (PSU)`, na.rm = TRUE)) %>%
  mutate(`Conductivity (µS/cm)` = median(`Conductivity (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Non-Linear Function (µS/cm)` = median(`Conductivity, Non-Linear Function (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Conductivity, Specific (µS/cm)` = median(`Conductivity, Specific (µS/cm)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen (mg/L)` = median(`Dissolved Oxygen (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Saturated (%)` = median(`Dissolved Oxygen, Saturated (%)`, na.rm = TRUE)) %>%
  mutate(`Dissolved Oxygen, Local (%)` = median(`Dissolved Oxygen, Local (%)`, na.rm = TRUE)) %>%
  mutate(`Oxygen Reduction Potential (mV)` = median(`Oxygen Reduction Potential (mV)`, na.rm = TRUE)) %>%
  mutate(`Chlorophyll (RFU)` = median(`Chlorophyll (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Algae (RFU)` = median(`Total Algae (RFU)`, na.rm = TRUE)) %>%
  mutate(`Total Dissolved Solids (mg/L)` = median(`Total Dissolved Solids (mg/L)`, na.rm = TRUE))%>%
  # take out columns we won't be using (pH has been renamed pH_YSI)
  select(-Time, -Date, -`Chlorophyll (µg/L)`, -`Total Algae (cells/L)`, -pH) %>%
  distinct() %>%
  ungroup() %>%
  rename(Date = date_year) %>%
  relocate(Date, .before = `Temperature (ºF)`) 

```


## Hach
### Daily
```{r data format: Hach daily values}
dfHach_daily <- dfHach %>%
  filter(Location == "Ocean") %>%
  select(-Location) %>%
  rename(pH_Hach = pH) %>%
  mutate(Date = as.Date(Date))

```


### Weekly Medians
```{r data format: Hach weekly medians}
dfHach_weekly <- dfHach_daily %>%
  # create year, month, week columns
  mutate(date_week = ceiling_date(as.Date(Date), "week")-1) %>%
  # group by year, month, week to take comparisons of each week within the time series
  group_by(date_week) %>%
  # take weekly medians since some data is taking 2x per week
  mutate(`Turbidity (RFU)` = median(`Turbidity (RFU)`, na.rm = TRUE)) %>%
  mutate(`Phosphate (mg/L)` = median(`Phosphate (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Nitrate, Mid-Range (mg/L)` = median(`Nitrate, Mid-Range (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Nitrate, High-Range (mg/L)` = median(`Nitrate, High-Range (mg/L)`, na.rm = TRUE)) %>%
  mutate(pH_Hach = median(pH_Hach, na.rm = TRUE)) %>%
  mutate(`Alkalinity (mg/L CaCO3)` = median(`Alkalinity (mg/L CaCO3)`, na.rm = TRUE)) %>%
  mutate(`Silica (mg/L)` = median(`Silica (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Iron (mg/L)` = median(`Iron (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Ammonia (mg/L)` = median(`Ammonia (mg/L)`, na.rm = TRUE)) %>%
  select(-Date) %>%
  distinct() %>%
  ungroup() %>%
  rename(Date = date_week) %>%
  relocate(Date, .before = `Turbidity (RFU)`) 

```

### Monthly Medians
```{r data format: Hach monthly medians}
dfHach_monthly <- dfHach_daily %>%
  # create column with Date of last day of each month of each year
  mutate(date_month = ceiling_date(as.Date(Date), "month")-1) %>%
  # group by year, month to take comparisons of each month within the time series
  group_by(date_month) %>%
  # take monthly medians
  mutate(`Turbidity (RFU)` = median(`Turbidity (RFU)`, na.rm = TRUE)) %>%
  mutate(`Phosphate (mg/L)` = median(`Phosphate (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Nitrate, Mid-Range (mg/L)` = median(`Nitrate, Mid-Range (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Nitrate, High-Range (mg/L)` = median(`Nitrate, High-Range (mg/L)`, na.rm = TRUE)) %>%
  mutate(pH_Hach = median(pH_Hach, na.rm = TRUE)) %>%
  mutate(`Alkalinity (mg/L CaCO3)` = median(`Alkalinity (mg/L CaCO3)`, na.rm = TRUE)) %>%
  mutate(`Silica (mg/L)` = median(`Silica (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Iron (mg/L)` = median(`Iron (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Ammonia (mg/L)` = median(`Ammonia (mg/L)`, na.rm = TRUE)) %>%
  select(-Date) %>%
  distinct() %>%
  ungroup() %>%
  rename(Date = date_month) %>%
  relocate(Date, .before = `Turbidity (RFU)`) 

```

### Yearly Medians
```{r data format: Hach yearly medians}
dfHach_yearly <- dfHach_daily %>%
  # create year column
  mutate(date_year = ceiling_date(as.Date(Date), "year")-1) %>%
  # group by year to take comparisons of each year within the time series
  group_by(date_year) %>%
  # take yearly medians
  mutate(`Turbidity (RFU)` = median(`Turbidity (RFU)`, na.rm = TRUE)) %>%
  mutate(`Phosphate (mg/L)` = median(`Phosphate (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Nitrate, Mid-Range (mg/L)` = median(`Nitrate, Mid-Range (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Nitrate, High-Range (mg/L)` = median(`Nitrate, High-Range (mg/L)`, na.rm = TRUE)) %>%
  mutate(pH_Hach = median(pH_Hach, na.rm = TRUE)) %>%
  mutate(`Alkalinity (mg/L CaCO3)` = median(`Alkalinity (mg/L CaCO3)`, na.rm = TRUE)) %>%
  mutate(`Silica (mg/L)` = median(`Silica (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Iron (mg/L)` = median(`Iron (mg/L)`, na.rm = TRUE)) %>%
  mutate(`Ammonia (mg/L)` = median(`Ammonia (mg/L)`, na.rm = TRUE)) %>%
  select(-Date) %>%
  distinct() %>%
  ungroup() %>%
  rename(Date = date_year) %>%
  relocate(Date, .before = `Turbidity (RFU)`) 


```


## Combine Data and Reformat
```{r data format: combine data}
# combine dfYSI_daily and dfHach_daily, with dfYSI_daily being first because it has more observations
dfYSIHach_daily <- left_join(dfYSI_daily, dfHach_daily, by = join_by(Date)) %>%
  # Add Time Interval column to designate these values as Daily
  mutate(`Time Interval` = "Daily") %>%
  relocate(`Time Interval`, .before = Date)

# combine dfYSI_weekly and dfHach_weekly, with dfYSI_weekly being first because it has more observations
dfYSIHach_weekly <- left_join(dfYSI_weekly, dfHach_weekly, by = join_by(Date)) %>%
   # Add Time Interval column to designate these values as Weekly
  mutate(`Time Interval` = "Weekly") %>%
  relocate(`Time Interval`, .before = Date)

# combine dfYSI_yearly and dfHach_yearly, with dfYSI_yearly being first because it has more observations
dfYSIHach_yearly <- left_join(dfYSI_yearly, dfHach_yearly, by = join_by(Date)) %>%
   # Add Time Interval column to designate these values as Yearly
  mutate(`Time Interval` = "Yearly") %>%
  relocate(`Time Interval`, .before = Date)

# combine dfYSI_monthly and dfHach_monthly, with dfYSI_monthly being first because it has more observations
dfYSIHach_monthly <- left_join(dfYSI_monthly, dfHach_monthly, by = join_by(Date))  %>%
   # Add Time Interval column to designate these values as Monthly
  mutate(`Time Interval` = "Monthly") %>%
  relocate(`Time Interval`, .before = Date)

# Combine all data into one dataframe
dfYSIHach_all <- rbind(dfYSIHach_daily, dfYSIHach_weekly, dfYSIHach_monthly, dfYSIHach_yearly)

```


# Write Data Files
```{r write csv file}
# create csv file of YSI and Hach daily values
write_csv(dfYSIHach_daily, 
          file = "OutputData/DescriptiveStatistics/HachYSI-Data-Medians-Daily.csv", 
          col_names = TRUE)

# create csv file of YSI and Hach weekly values
write_csv(dfYSIHach_weekly,
          file = "OutputData/DescriptiveStatistics/HachYSI-Data-Medians-Weeky.csv",
          col_names = TRUE)

# create csv file of YSI and Hach monthly values
write_csv(dfYSIHach_monthly,
          file = "OutputData/DescriptiveStatistics/HachYSI-Data-Medians-Monthly.csv",
          col_names = TRUE)

# create csv file of YSI and Hach yearly values
write_csv(dfYSIHach_yearly,
          file = "OutputData/DescriptiveStatistics/HachYSI-Data-Medians-Yearly.csv",
          col_names = TRUE)

# create csv file of YSI and Hach all values
write_csv(dfYSIHach_all,
          file = "OutputData/DescriptiveStatistics/HachYSI-Data-Medians-All.csv",
          col_names = TRUE)

```

