---
title: "Ocean Data Wrangling: 2022-Present"
Collaborators: Renee Grambihler
Project: OceanWaterQuality
Institution: Biosphere 2 Ocean
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Ocean Data Wrangling: 2022-Present

```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
library(here)
# turns off scientific notation
options(scipen = 999)
```

## Import Data

### Old B2 SCADA Database: 2011-2022

Use `01-YSI-Data-Clean-2011-2022.csv` created in `01-YSI-DataWrangling-2011-2022.Rmd`

```{r import-data-old-scada-database}
dfHistorical <- read_csv("InputData/Clean/01-DataWrangling/YSI-Data-Clean-2011-2022.csv") 
```

## New B2 SCADA Database: 1/22/22 - Present

These files are pulled from the [B2 SCADA Database](<https://data.b2.arizona.edu/Bio2Controls/OCDataAgreement.jsp>). Each file will download with a numeric identifier at the end of the file name. Delete this before running the files through this code. I might figure out how to get around that via code one day, but not today.

```{r SCADA-files-from-1/22/22-present-path}
# file path to folder containing new scada database 1/22/22-09/25/2024 ocean water quality files
path <- "Data/Raw/YSI/SCADA/SCADA_2022_2024"
```

```{r import-SCADA-files-from-1/22/22-present}

# chlorophyll in RFU from 1/22/22-present from current YSI EXO3 Sonde
exo_chlorophyll <- read_csv(here(path,
                            "UAB2_BIO1_B4000_YSI_EXO3_CHLOROPHYLL_RFU.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Chlorophyll (RFU)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time)) 

# electrical conductivity in uS/cm from 1/22/22-present from current YSI EXO3 Sonde
exo_conductivity <- read_csv(here(path, 
                                    "UAB2_BIO1_B4000_YSI_EXO3_COND_USPCM.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Conductivity (µS/cm)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# nLF conductivity in uS/cm from 1/22/22-present from current YSI EXO3 Sonde
exo_nlf_conductivity <- read_csv(here(path, 
                                "UAB2_BIO1_B4000_YSI_EXO3_NLF_COND_USPCM.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Conductivity, Non-Linear Function (µS/cm)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# dissolved oxygen in % local from 1/22/22-present from current YSI EXO3 Sonde
exo_do_local <- read_csv(here(path, 
                                    "UAB2_BIO1_B4000_YSI_EXO3_ODO__LOCAL.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen, Local (%)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# dissolved oxygen in % saturated from 1/22/22-present from current YSI EXO3 Sonde
exo_do_saturated <- read_csv(here(path, 
                                    "UAB2_BIO1_B4000_YSI_EXO3_ODO__SAT.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen, Saturated (%)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# dissolved oxygen in mg/L from 1/22/22-present from current YSI EXO3 Sonde
exo_do <- read_csv(here(path, "UAB2_BIO1_B4000_YSI_EXO3_ODO_MGPL.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Dissolved Oxygen (mg/L)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# reduction potential (ORP) in mV from 1/22/22-present from current YSI EXO3 Sonde
exo_orp <- read_csv(here(path, "UAB2_BIO1_B4000_YSI_EXO3_ORP_MV.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Oxygen Reduction Potential (mV)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# pH in pH units from 1/22/22-present from current YSI EXO3 Sonde
exo_ph <- read_csv(here(path, "UAB2_BIO1_B4000_YSI_EXO3_PH.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         pH = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# pH in mV from 1/22/22-present from current YSI EXO3 Sonde
exo_ph_mv <- read_csv(here(path, "UAB2_BIO1_B4000_YSI_EXO3_PH_MV.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `pH (mV)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# salinity in psu from 1/22/22-present from current YSI EXO3 Sonde
exo_salinity <- read_csv(here(path,
                                    "UAB2_BIO1_B4000_YSI_EXO3_SAL_PSU.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Salinity (PSU)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# specific conductivity in uS/cm from 1/22/22-present from current YSI EXO3 Sonde
exo_sp_conductivity <- read_csv(here(path, 
                                  "UAB2_BIO1_B4000_YSI_EXO3_SPCOND_USPCM.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Conductivity, Specific (µS/cm)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# total algae phycoerythrin in RFU from 1/22/22-present from current YSI EXO3 Sonde
exo_total_algae <- read_csv(here(path, 
                                  "UAB2_BIO1_B4000_YSI_EXO3_TAL_PE_RFU.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Total Algae (RFU)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# total dissolved solids in mg/L from 1/22/22-present from current YSI EXO3 Sonde
exo_total_dissolved_solids <- read_csv(here(path,
                                    "UAB2_BIO1_B4000_YSI_EXO3_TDS_MGPL.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Total Dissolved Solids (mg/L)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# temperature in degrees Celcius from 1/22/22-present from current YSI EXO3 Sonde
exo_temp_c <- read_csv(here(path, "UAB2_BIO1_B4000_YSI_EXO3_TEMP_C.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Temperature (ºC)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

# temperature in degrees Fahrenheit from 1/22/22-present from current YSI EXO3 Sonde
exo_temp_f <- read_csv(here(path, "UAB2_BIO1_B4000_YSI_EXO3_TEMP_F.csv"),
                              skip = 2,
                              col_select = c("TIMESTAMP", "VALUE")) %>%
  rename(datetime = TIMESTAMP,
         `Temperature (ºF)` = VALUE) %>%
  # separate datetime into Date and Time columns
  separate(col = datetime, into = c("Date", "Time"), sep = " ") %>%
  # make sure only data recorded on minutes 00, 15, 30, and 45 are kept
  filter(grepl(':00:|:15:|:30:|:45:', Time))

```

## ProDSS Backup

This file is from the YSI ProDSS instrument. The EXO3 was sent in for maintenance from 11/17/2023 - 1/10/2024, so it was not recording the to the database. The YSI ProDSS was used to record data while the EXO3 was away. Unfortunately, not all of the same variables were recorded, but this will be fixed if the EXO3 ever needs to go back.

```{r import-ProDSS-file}
dfProDSS <- read_csv("Data/Raw/YSI/ProDSS/prodssbackup-231117-240112.csv",
                     # files downloaded from the YSI instruments are encoded as UTF-16, but most code languages only read UTF-8. Make sure to keep the below line in to fix that.
                     locale=locale(encoding="UTF-16"),
                     # first line says sep=, so that excel reads in the csv file like an excel file so make sure to skip that. can only see this line when viewing with a text editor
                     skip = 1) %>%
  # remove columns we won't need
  select(-`User ID`, -`Unit ID`, -Site)%>%
  rename(`Temperature (ºF)` = "°F",
         `Dissolved Oxygen, Local (%)` = "DO %",
         `Dissolved Oxygen (mg/L)` = "DO mg/L",
         `Conductivity (µS/cm)` = "C-uS/cm",
         `Salinity (PSU)` = "SAL-PSU",
         `Oxygen Reduction Potential (mV)` = "ORP mV",
         `Chlorophyll (RFU)` = "Chl RFU",
         `Total Algae (RFU)` = "PE RFU") 

```

## Data Wrangling

Goal here is to make all files look the same so they can be combined later

### Historical File

```{r data-wrangling-historical}
dfHistorical <- dfHistorical %>%
  relocate(`pH (mV)`, .after = pH) %>%
  filter(Date < "2021-05-27")
  
```

Check Data:

```{r dfhistorical-data-check}
dfHistorical %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows
```

### New Database Files

Combine all files from new database and make the resulting dataframe look like dfHistorical. Chop this dataframe in half so that we can sandwich in the ProDSS dataframe.

```{r data-wrangling-new-database}
# create new database YSI EXO3 dataframe by joining the files for each variable from this instrument based on the datetime column
dfNewDB <- exo_temp_f %>%
  left_join(exo_temp_c, by = join_by(Date, Time)) %>%
  left_join(exo_ph, by = join_by(Date, Time)) %>%
  left_join(exo_ph_mv, by = join_by(Date, Time)) %>%
  left_join(exo_salinity, by = join_by(Date, Time)) %>%
  left_join(exo_conductivity, by = join_by(Date, Time)) %>%
  left_join(exo_orp, by = join_by(Date, Time)) %>%
  left_join(exo_total_algae, by = join_by(Date, Time)) %>%
  mutate(`Chlorophyll (µg/L)` = NA) %>%
  mutate(`Total Algae (cells/L)` = NA) %>%
  left_join(exo_do_saturated, by = join_by(Date, Time)) %>%
  left_join(exo_do, by = join_by(Date, Time)) %>%
  left_join(exo_do_local, by = join_by(Date, Time)) %>%
  left_join(exo_nlf_conductivity, by = join_by(Date, Time)) %>%
  left_join(exo_sp_conductivity, by = join_by(Date, Time)) %>%
  left_join(exo_total_dissolved_solids, by = join_by(Date, Time)) %>%
  left_join(exo_chlorophyll, by = join_by(Date, Time)) %>%
  mutate(Date = as.Date(Date))

# df of data recorded before EXO3 got sent to be fixed
dfNewDB_beginning <- dfNewDB %>%
  filter(Date < "2023-11-17" & Time < "14:30:00")

# df of data recorded after EXO3 came back from being fixed
dfNewDB_end <- dfNewDB %>%
  filter(Date > "2024-01-10" & Time < "15:16:00")

```

Check Data:

```{r newdb-data-check}
dfNewDB %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

dfNewDB_beginning %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

dfNewDB_end %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

```

### ProDSS File

Make dfProDSS look like dfHistorical

```{r data-wrangle-ProDSS}
# add NA columns with same names as columns missing from other dataframes
dfProDSSEdit <- dfProDSS %>%
  mutate(`Temperature (ºC)` = NA,
         `pH (mV)` = NA,
         `Conductivity, Non-Linear Function (µS/cm)` = NA,
         `Conductivity, Specific (µS/cm)` = NA,
         `Dissolved Oxygen, Saturated (%)` = NA,
         `Chlorophyll (µg/L)` = NA,
         `Total Algae (cells/L)` = NA,
         `Total Dissolved Solids (mg/L)` = NA) %>%
  # move all columns to same places as other dataframes
  relocate(`Temperature (ºC)`, .after = `Temperature (ºF)`) %>%
  relocate(pH, .after = `Temperature (ºC)`) %>%
  relocate(`pH (mV)`, .after = pH) %>%
  relocate(`Salinity (PSU)`, .after = `pH (mV)`) %>%
  relocate(`Conductivity (µS/cm)`, .after = `Salinity (PSU)`) %>%
  relocate(`Oxygen Reduction Potential (mV)`, .after = `Conductivity (µS/cm)`) %>%
  relocate(`Total Algae (RFU)`, .after = `Oxygen Reduction Potential (mV)`) %>%
  relocate(`Chlorophyll (µg/L)`, .after = `Total Algae (RFU)`) %>%
  relocate(`Total Algae (cells/L)`, .after = `Chlorophyll (µg/L)`) %>%
  relocate(`Dissolved Oxygen, Saturated (%)`, .after = `Total Algae (cells/L)`) %>%
  relocate(`Dissolved Oxygen (mg/L)`, .after = `Dissolved Oxygen, Saturated (%)`) %>%
  relocate(`Dissolved Oxygen, Local (%)`, .after = `Dissolved Oxygen (mg/L)`) %>%
  relocate(`Conductivity, Non-Linear Function (µS/cm)`, .after = `Dissolved Oxygen, Local (%)`) %>%
  relocate(`Conductivity, Specific (µS/cm)`, .after = `Conductivity, Non-Linear Function (µS/cm)`) %>%
  relocate(`Chlorophyll (RFU)`, .after = `Conductivity, Specific (µS/cm)`) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>%
  # filter out extra rows that got added when file was downloaded off of instrument
  filter(Date <= "2024-01-10" & Time <= "15:15:00")

```

Data Check:

```{r ProDSS-data-check}
dfProDSSEdit %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows
```

# Combine Files

Combine all dataframes in a way that gives us all of the data and separate datetime column into Date and Time so it's more usable in later analyses.

```{r combine-files}
dfOceanWaterQuality <- rbind(dfHistorical,
                             dfNewDB_beginning,
                             dfProDSSEdit,
                             dfNewDB_end) %>%
  distinct() %>%
  # make sure all columns are proper data type
  mutate(Date = ymd(Date),
         `Temperature (ºF)` = as.double(`Temperature (ºF)`),
         `Temperature (ºC)` = as.double(`Temperature (ºC)`),
         pH = as.double(pH),
         `pH (mV)` = as.double(`pH (mV)`),
         `Salinity (PSU)` = as.double(`Salinity (PSU)`),
         `Conductivity (µS/cm)` = as.double(`Conductivity (µS/cm)`),
         `Conductivity, Non-Linear Function (µS/cm)` = as.double(`Conductivity, Non-Linear Function (µS/cm)`),
         `Conductivity, Specific (µS/cm)` = as.double(`Conductivity, Specific (µS/cm)`),
         `Dissolved Oxygen (mg/L)` = as.double(`Dissolved Oxygen (mg/L)`),
         `Dissolved Oxygen, Saturated (%)` = as.double(`Dissolved Oxygen, Saturated (%)`),
         `Dissolved Oxygen, Local (%)` = as.double(`Dissolved Oxygen, Local (%)`),
         `Oxygen Reduction Potential (mV)` = as.double(`Oxygen Reduction Potential (mV)`),
         `Chlorophyll (RFU)` = as.double(`Chlorophyll (RFU)`),
         `Chlorophyll (µg/L)` = as.double(`Chlorophyll (µg/L)`),
         `Total Algae (RFU)` = as.double(`Total Algae (RFU)`),
         `Total Algae (cells/L)` = as.double(`Total Algae (cells/L)`),
         `Total Dissolved Solids (mg/L)` = as.double(`Total Dissolved Solids (mg/L)`)) %>%
  filter(grepl(':00:|:15:|:30:|:45:', Time))
  

```

Final data check

```{r check-data-final}
dfOceanWaterQuality %>%
  unique() %>% # keep only unique rows
  group_by(Date) %>% # group each date together
  summarise(n = n()) %>% # count the number of rows per group (date)
  filter(n > 96) # keep only groups with more than 96 rows

```

# Write File

```{r write-data-to-file}
write_csv(dfOceanWaterQuality,
          file = "Data/Clean/01-DataWrangling/YSI-Data-Clean-All.csv",
          col_names = TRUE)
```
