---
title: "HACH Data Wrangling: 2017-2024"
Collaborators: Renee Grambihler
Project: "The Biosphere 2 Ocean mesocosm: (re)engineered for state-of-the-art coral reef resilience experiments"
Institution: Biosphere 2 Reef Lab
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
# Hach Data Wrangling: 2017-2024
## Input and Output Files
### Input Files
InputData/Raw/Hach/:  
  `B2ONutrientData.xlsx`: all Biosphere 2 Systems nutrient water quality data obtained via colorimetry and titration (Hach brand); systems include Ocean, Mangroves, and 2 Coral Raceways; sheet "Pre-formatted Data" holds data from 2017-2022; sheet "Data" holds data from 2022-2024 that is formatted more for analysis in R.  
    - `Pre-formatted Data`: variable names are repeated for each location and made distinct in file by color coding, key is found in Notes column but Ocean is column 2:11, Mangroves is 12:19, Raceway 1 is 20:29, and Raceway 2 is 30:39  
        - variables:  
          `Date`: date in mm/dd/yyyy  
          `Turbidity`: numeric; Ocean turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
          `Phosphate`: numeric; Ocean phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
          `Nitrate MR`: numeric; Ocean nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
          `Nitrate HR`: numeric; Ocean nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
          `pH`: numeric; Ocean pH in pH units  
          `Alkalinity`: numeric; Ocean calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
          `Silica`: numeric; Ocean silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
          `Iron`: numeric; Ocean iron concentration in range 0.009-1.400 measured in mg/L  
          `Ammonia`: numeric; Ocean ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  
          `ORP`: numeric; Ocean oxidation-reduction potential (ORP) measured in miliVolts  
          `Turbidity`: numeric; Mangroves turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
          `Phosphate`: numeric; Mangroves phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
          `Nitrate MR`: numeric; Mangroves nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
          `Nitrate HR`: numeric; Mangroves nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
          `pH`: numeric; Mangroves pH in pH units  
          `Alkalinity`: numeric; Mangroves calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
          `Silica`: numeric; Mangroves silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
          `ORP`: numeric; Mangroves oxidation-reduction potential (ORP) measured in miliVolts  
          `Turbidity`: numeric; Raceway 1 turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
          `Phosphate`: numeric; Raceway 1 phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
          `Nitrate MR`: numeric; Raceway 1 nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
          `Nitrate HR`: numeric; Raceway 1 nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
          `pH`: numeric; Raceway 1 pH in pH units  
          `Alkalinity`: numeric; Raceway 1 calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
          `Silica`: numeric; Raceway 1 silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
          `Iron`: numeric; Raceway 1 iron concentration in range 0.009-1.400 measured in mg/L  
          `Ammonia`: numeric; Raceway 1 ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  
          `ORP`: numeric; Raceway 1 oxidation-reduction potential (ORP) measured in miliVolts  
          `Turbidity`: numeric; Raceway 2 turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
          `Phosphate`: numeric; Raceway 2 phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
          `Nitrate MR`: numeric; Raceway 2 nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
          `Nitrate HR`: numeric; Raceway 2 nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
          `pH`: numeric; Raceway 2 pH in pH units  
          `Alkalinity`: numeric; Raceway 2 calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
          `Silica`: numeric; Raceway 2 silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
          `Iron`: numeric; Raceway 2 iron concentration in range 0.009-1.400 measured in mg/L  
          `Ammonia`: numeric; Raceway 2 ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  
          `Initials`: character; initials of person or people taking measurements  
          `Notes`: character; key for color coding, any notes from that measurement  
        
    - `Data`: 
        - variables:  
          `Date`: date in mm/dd/yyy
          `Location`: character; location of measurements; Ocean, Mangroves, Raceway 1, or Raceway 2  
          `Turbidity`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
          `Phosphate`: numeric; phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
          `Nitrate MR`: numeric; nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
          `Nitrate HR`: numeric; nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
          `pH`: numeric; pH in pH units  
          `Alkalinity`: calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
          `Silica`: numeric; silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
          `Iron`: numeric; iron concentration in range 0.009-1.400 measured in mg/L  
          `Ammonia`: numeric; ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  
          `Initials`: character; initials of person or people taking measurements  
          `Notes`: character; any notes from that measurement  

### Output Files
InputData/Clean/01-DataWrangling/:  
  `Hach-Data-Clean.csv` 
    - variables: 
      `Date`: date in mm/dd/yy  
      `Location`: character; location of measurements; Ocean, Mangroves, Raceway 1, or Raceway 2  
      `Turbidity (FAU)`: numeric; turbidity (water transparency) in the range 21-1000 Formazine Attenuation Units (FAU)  
      `Phosphate (mg/L)`: numeric; phosphate (PO4 3-) concentration in range 0.02-2.50 measured in mg/L  
      `Nitrate, Mid-Range (mg/L)`: numeric; nitrate (NO3-) mid-range concentration from 0.2-5.0 measured in mg/L  
      `Nitrate, High-Range (mg/L)`: numeric; nitrate (NO3-) high-range from 0.3-30.0 measured in mg/L  
      `pH`: pH in pH units  
      `Alkalinity (mg/L CaCO3)`: numeric; calcium carbonate (CaCO3) concentration in range 100-400 measured in mg/L  
      `Silica (mg/L)`: numeric; silica (SiO2) concentration in range 0.01-1.60 measured in mg/L  
      `Iron (mg/L)`: numeric; iron concentration in range 0.009-1.400 measured in mg/L  
      `Ammonia (mg/L)`: numeric; ammonia (NH3) concentration in range 0.01-0.50 measured in mg/L  
      `ORP`:  numeric;oxidation-reduction potential (ORP) measured in miliVolts  
      `Initials`: character; initials of person or people taking measurements  
      `Notes`: character; any notes from that measurement  


```{r label='Import-Libraries'}
library(tidyverse)
library(readxl)
library(lubridate)
# remove scientific notation
options(scipen = 999)
```

# Import Data
```{r import data}
# internal use: download B20NutrientData excel file from the Box. The Pre-formatted Data (2017-2022) is on one sheet, and the Formatted Data (2022-2024) is on a different sheet. 

# Import Pre-Formatted Data sheet from excel file (sheet "Pre-formatted Data")
PreFormattedData <- read_excel("InputData/Raw/Hach/B2ONutrientData.xlsx",
                               sheet = "Pre-formatted Data",
                               col_names = TRUE)

# Import Formatted Data sheet from excel file (sheet "Data")
FormattedData <- read_excel("InputData/Raw/Hach/B2ONutrientData.xlsx", 
                            sheet = "Data", 
                            col_names = TRUE)
```

# Data Tidying
## Pre-Formatted Data
```{r data tidying: Pre-Formatted Data sheet}
# get rid of (...##) at end of column names
for (col in 1:ncol(PreFormattedData)){
    colnames(PreFormattedData)[col] <-  gsub("\\.*[0123456789]", "", colnames(PreFormattedData)[col])
    }

# separating large dataframe into separate ones for each testing location (Ocean, Mangrove, Raceway 1, and Raceway 2) in order to add a Location column 
OceanFrame <- PreFormattedData[, c(1:11, 39)]%>%
  mutate(Location = "Ocean")

MangroveFrame <- PreFormattedData[, c(1, 12:19, 39)] %>%
  mutate(Location = "Mangroves",
         Alkalinity = NA, 
         Silica = NA, 
         Iron = NA, 
         Ammonia = NA)
  
RacewayFrame1 <- PreFormattedData[, c(1, 20:29, 39)] %>%
  mutate(Location = "Raceway 1")

RacewayFrame2 <- PreFormattedData[, c(1, 30:39)] %>%
  mutate(Location = "Raceway 2",
         ORP = NA)

# combine the dataframes by rows to get longer format data
PreFormattedDataTidy <- rbind(OceanFrame, MangroveFrame, RacewayFrame1, RacewayFrame2) %>%
  mutate(Silica = as.numeric(Silica),
         `Nitrate MR` = as.numeric(`Nitrate MR`),
         `Nitrate HR` = as.numeric(`Nitrate HR`),
         pH = as.numeric(pH),
         Date = format(Date, "%m/%d/%y"), 
         Notes = NA) %>%
  relocate(Location, .after = Date)

```

## Formatted Data
```{r data tidying: Data sheet}
# changing --- and +++ values to the lowest (---) and highest (+++) value for the range read by each nutrient test (see metadata for full explanation of each range). 
FormattedDataTidy <- FormattedData %>%
  mutate(Turbidity = round(Turbidity, digits = 1),
         Phosphate = ifelse(Phosphate == "---", 0.02, Phosphate),
         Phosphate = ifelse(Phosphate == "+++", 2.50, Phosphate),
         Phosphate = as.numeric(Phosphate),
         Phosphate = round(Phosphate, digits = 2),
         `Nitrate MR` = ifelse(`Nitrate MR` == "---", 0.2, `Nitrate MR`),
         `Nitrate MR` = ifelse(`Nitrate MR` == "+++", 5.0, `Nitrate MR`),
         `Nitrate MR` = as.numeric(`Nitrate MR`),
         `Nitrate MR` = round(`Nitrate MR`, digits = 2),
         `Nitrate HR` = ifelse(`Nitrate HR` == "---", 0.3, `Nitrate HR`),
         `Nitrate HR` = ifelse(`Nitrate HR` == "+++", 30.0, `Nitrate HR`),
         `Nitrate HR` = as.numeric(`Nitrate HR`),
         `Nitrate HR` = round(`Nitrate HR`, digits = 2),
         pH = round(pH, digits = 2),
         Alkalinity = round(Alkalinity, digits = 1),
         Silica = ifelse(Silica == "---", 0.01, Silica),
         Silica = ifelse(Silica == "+++", 1.60, Silica),
         Silica = round(as.numeric(Silica), digits = 2),
         Iron = ifelse(Iron == "---", 0.009, Iron),
         Iron = ifelse(Iron == "+++", 1.400, Iron),
         Iron = round(Iron, digits = 3),
         Ammonia = ifelse(Ammonia == "---", 0.01, Ammonia),
         Ammonia = ifelse(Ammonia == "+++", 0.50, Ammonia),
         Ammonia = as.numeric(Ammonia),
         Ammonia = round(Ammonia, digits = 2),
         Date = format(Date, "%m/%d/%y"),
         ORP = NA) %>%
  relocate(ORP, .after = Ammonia) %>%
  select(-`...14`)

```

## Combine Data
```{r Combine data from Pre-formatted and Data sheets}
AllData <- rbind(PreFormattedDataTidy, FormattedDataTidy) %>%
  mutate(Turbidity = round(Turbidity, digits = 1),
         Phosphate = ifelse(Phosphate == "---", 0.02, Phosphate),
         Phosphate = ifelse(Phosphate == "+++", 2.50, Phosphate),
         Phosphate = as.numeric(Phosphate),
         Phosphate = round(Phosphate, digits = 2),
         `Nitrate MR` = ifelse(`Nitrate MR` == "---", 0.2, `Nitrate MR`),
         `Nitrate MR` = ifelse(`Nitrate MR` == "+++", 5.0, `Nitrate MR`),
         `Nitrate MR` = as.numeric(`Nitrate MR`),
         `Nitrate MR` = round(`Nitrate MR`, digits = 2),
         `Nitrate HR` = ifelse(`Nitrate HR` == "---", 0.3, `Nitrate HR`),
         `Nitrate HR` = ifelse(`Nitrate HR` == "+++", 30.0, `Nitrate HR`),
         `Nitrate HR` = as.numeric(`Nitrate HR`),
         `Nitrate HR` = round(`Nitrate HR`, digits = 2),
         pH = round(pH, digits = 2),
         Alkalinity = round(Alkalinity, digits = 1),
         Silica = ifelse(Silica == "---", 0.01, Silica),
         Silica = ifelse(Silica == "+++", 1.60, Silica),
         Silica = round(as.numeric(Silica), digits = 2),
         Iron = ifelse(Iron == "---", 0.009, Iron),
         Iron = ifelse(Iron == "+++", 1.400, Iron),
         Iron = round(Iron, digits = 3),
         Ammonia = ifelse(Ammonia == "---", 0.01, Ammonia),
         Ammonia = ifelse(Ammonia == "+++", 0.50, Ammonia),
         Ammonia = as.numeric(Ammonia),
         Ammonia = round(Ammonia, digits = 2),
         Date = as.Date(Date, format = "%m/%d/%y")) %>%
  rename(`Turbidity (FAU)` = Turbidity,
         `Phosphate (mg/L)` = Phosphate,
         `Nitrate, Mid-Range (mg/L)` = `Nitrate MR`,
         `Nitrate, High-Range (mg/L)` = `Nitrate HR`,
         `Alkalinity (mg/L CaCO3)` = Alkalinity,
         `Silica (mg/L)` = Silica,
         `Iron (mg/L)` = Iron,
         `Ammonia (mg/L)` = Ammonia)

```

## Save Data
```{r save data}
write_csv(AllData,
          file = "InputData/Clean/01-DataWrangling/Hach-Data-Clean.csv")

```

