---
title: "World Oceans Data Wrangling"
Collaborators: Renee Grambihler
Project: "The Biosphere 2 Ocean mesocosm: (re)engineered for state-of-the-art coral reef resilience experiments"
Institution: Biosphere 2 Reef Lab
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
# World Oceans Data Wrangling
## Input and Output Files
### Input Files
InputData/Clean/01-DataWrangling: 
  `WorldOceans-Data-Clean.csv`: water quality data from Atlantic and Pacific Oceans from 2011-2024
    - variables:  
      `Date`: date; mm/dd/yy 
      `Ocean`: character; name of ocean data was obtained from  
      `Region`: character; region of ocean that data was obtained from  
      `Depth (m)`: numeric; depth underwater data was obtained from  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `Salinity (PSU)`: numeric; salinity in Potential Salinity Units  
      `Phosphate (µmol/kg)`: numeric; phosphate concentration in µmol/kg  
      `Silicate (µmol/kg)`: numeric; silicate concentration in µmol/kg    
      `Nitrate (µmol/kg)`: numeric; nitrate concentration in µmol/kg  
      `pH`: numeric; pH in pH units  
      `Chlorophyll (µg/L)`: numeric; chlorophyll concentration in µg/L   
      `Alkalinity (µEq/L)`: numeric; alkalinity concentration in µEquivalents/L  
      `Dissolved Organic Carbon (µmol/kg)`: numeric; dissolved organic carbon concentration in µmol/kg  
      `Iron (mg/L)`: numeric; iron concentration in mg/L  
  
  `WorldOceans-RegionsCoordinates-Clean.csv`: ocean regions found in `WorldOceans-Data-Clean.csv` and their coordinates in longitude and latitude 
    - variables:  
      `Region`: character; ocean region corresponding to those found in `WorldOceans-Data-Clean.csv`  
      `Latitude`: numeric; latitude of region datum point  
      `Longitude`: numeric; longitude of region datum point  
  

### Output Files
InputData/Clean/01-DataWrangling: 
  `WorldOceans-Data-Clean-Final.csv`: combination of `WorldOceans-RegionsCoordinates-Clean.csv` and `WorldOceans-Data-Clean.csv` with additional Location column for more specific geographic location label
    - variables:  
      `Date`: date; mm/dd/yy 
      `Ocean`: character; name of ocean data was obtained from  
      `Region`: character; region of ocean that data was obtained from  
      `Location`: character; more specific geographic location than region, such as an island, state, atoll, or reef  
      `Depth (m)`: numeric; depth underwater data was obtained from  
      `Temperature (ºC)`: numeric; temperature in degrees Celcius  
      `Salinity (PSU)`: numeric; salinity in Potential Salinity Units  
      `Phosphate (µmol/kg)`: numeric; phosphate concentration in µmol/kg  
      `Silicate (µmol/kg)`: numeric; silicate concentration in µmol/kg    
      `Nitrate (µmol/kg)`: numeric; nitrate concentration in µmol/kg  
      `pH`: numeric; pH in pH units  
      `Chlorophyll (µg/L)`: numeric; chlorophyll concentration in µg/L   
      `Alkalinity (µEq/L)`: numeric; alkalinity concentration in µEquivalents/L  
      `Dissolved Organic Carbon (µmol/kg)`: numeric; dissolved organic carbon concentration in µmol/kg  
      `Iron (mg/L)`: numeric; iron concentration in mg/L
      `Latitude`: numeric; latitude of region datum point  
      `Longitude`: numeric; longitude of region datum point


```{r label='Import-Libraries'}
library(tidyverse)
library(lubridate)
```

# Import Data
```{r import-data}
# `WorldOceans-Data-Clean.csv` was compiled separately in MS Excel from the data files in InputData/Raw/WorldOceans which was obtained from the NOAA National Centers for Environmental Information NCEI World Ocean Database
dfWOD <- read_csv("InputData/Clean/01-Wrangling/WorldOceans-Data-Clean.csv") 

# `WorldOceans-RegionsCoordinates-Clean.csv` was compiled in MS Excel using coordiantes of the locations in `WorldOceans-Data-Clean.csv`
dfCoords <- read_csv("InputData/Clean/01-Wrangling/WorldOceans-RegionsCoordinates-Clean.csv")

```

# Data Wrangling
## Traditional Wrangling
```{r data-wrangling}
# do some basic editing: date format, add in iron value that doesn't wanna import
dfWOD_edit <- dfWOD %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%y")) %>%
  mutate(`Iron (mg/L)` = ifelse(`Silicate (µmol/kg)` == "0.21", "0.024", `Iron (mg/L)`)) %>%
  # not sure where Indian Ocean data is from so taking it out
  filter(Ocean != "Indian")

# make dataframe of just Diane Marshall Islands data because it's in a different format than rest of data
dfMarshalls <- dfWOD_edit %>%
  filter(Date == "2024-08-29")
# change column names to correct units
colnames(dfMarshalls) <- c("Date", "Ocean", "Region", "Location", "Depth (m)", "Temperature (ºC)",
                           "Salinity (PSU)", "Phosphate (mg/L)", "Silicate (mg/L)", "Nitrate (mg/L)",
                           "pH", "Chlorophyll (µg/L)", "Alkalinity (mg/L)", 
                           "Dissolved Organic Carbon (µmol/kg)", "Iron (mg/L)")


# not sure why, but 01-Clean-WorldOceanData.csv has alkalinity as µEq/L even though only the raw data file from the Atlantic uses µEq/L. The other 2 (both in Pacific) both use µmol/kg. Just going to change both to mg/L now instead of with the rest of the nutrients later 

# atlantic ocean: uEq/L -> mg/L: uEq/L / 0.01998 / 1000 mEq/L = mg/L 
# not sure how exactly we get mg from mEq, but the 0.01998 is some conversion factor from USGS [Atkinson 1999]
dfWOD_atlantic <- dfWOD_edit %>%
  filter(Ocean == "Atlantic") %>%
  mutate(`Alkalinity (mg/L)` = `Alkalinity (µEq/L)` / 0.01998 / 1000, 
         .after = `Chlorophyll (µg/L)`) %>%
  select(-`Alkalinity (µEq/L)`)

# pacific ocean: µmol/kg -> mg/L: umol/L * density seawater (kg/L) * molecular mass CaCO3 (g/mol) / 1000000 umol/mol * 1000 mg/g
# = umol/L * 1.025 kg/L * 100.09 g/mol / 1000000 umol/mol * 1000 mg/g
dfWod_pacific <- dfWOD_edit %>%
  filter(Ocean == "Pacific") %>%
  # take out marshall islands since they are in dfMarshalls and in mg/L already 
  filter(Date != "2024-08-29") %>%
  mutate(`Alkalinity (mg/L)` = `Alkalinity (µEq/L)` * 1.025 * 100.09 / 1000000 * 1000, 
         .after = `Chlorophyll (µg/L)`) %>%
  select(-`Alkalinity (µEq/L)`)

# convert data to units that will match B2 data
dfWOD_correct <- rbind(dfWod_pacific, dfWOD_atlantic) %>%
  # make sure Marshall's data isn't in there
  filter(Date < "2024-08-29") %>%
  # convert from µmol/kg of each compound into mg/L using average density of seawater at sealevel (1.025 kg/L) and the molecular weight of each compound (ex: phosphate = [PO4]3- = 94.973 g/mol)
  # for alkalinity, convert uEq/L to mg/L using conversion factor 0.01998 (USGS): 
  #100 mg/L is 1000 umol/kg whis is 2 meq/L
  mutate(`Phosphate (mg/L)` = `Phosphate (µmol/kg)` * 1.025 * 94.973 / 1000000000,
         `Silicate (mg/L)` = `Silicate (µmol/kg)` * 1.025 * 76.08 / 1000000000,
         `Nitrate (mg/L)` = `Nitrate (µmol/kg)` * 1.025 * 62.0049 / 1000000000) %>%
  select(-`Phosphate (µmol/kg)`, -`Silicate (µmol/kg)`, -`Nitrate (µmol/kg)`) %>%
  relocate(contains(c("Phosphate (mg/L)", "Silicate (mg/L)", "Nitrate (mg/L)")), .after = `Salinity (PSU)`)

# combine dataframes
dfWOD_combo <- rbind(dfWOD_correct, dfMarshalls)

```

## Adding Info
```{r add-countries}
dfWOD_final <- dfWOD_combo %>%
  mutate(Latitude = NA,
         Longitude = NA)

locations <- unique(dfWOD_combo$Location)

for (i in locations) {
 dfWOD_final <- dfWOD_final %>%
   mutate(Latitude = ifelse(Location == i, dfCoords[dfCoords$Region == i, 2], Latitude),
          Longitude = ifelse(Location == i, dfCoords[dfCoords$Region == i, 3], Longitude))
}

dfWOD_final <- dfWOD_final %>%
  mutate(`Iron (mg/L)` = as.numeric(`Iron (mg/L)`),
         Latitude = as.numeric(Latitude),
         Longitude = as.numeric(Longitude))

```


# Write Data
```{r write-data}
write_csv(dfWOD_final,
          file = "Data/Clean/WorldOceans/WorldOceans-Data-Clean-Final.csv")
```

